---
title: "Total Abundance - Meta Analysis "
subtitle: "Analyzing the effect of fragmentation on total abundance from capture data and how this is influenced by landscape variables"
author: "Dillan Hoyt"
execute: 
  eval: true
  echo: false
  message: false
editor: visual
toc: true
toc-depth: 3
toc-title: Contents
---
### In this file, we are analyzing the total abundance data. The structure of the code is similar to the species richness file, but with some adjustments for the different response variable and potentially different covariates.

```{r}
libs <- c(
  "brms", "projpred", "tidybayes", "bayesplot", "emmeans",
  "posterior", "performance", "loo", "car", "dplyr",
  "tidyr", "tibble", "forcats", "ggplot2", "ggdist",
  "ggcorrplot", "scales", "patchwork", "corrr"
)

invisible(lapply(libs, library, character.only = TRUE))

#insert csv:
meta_AB <- read.csv("effect_size_totalabundance.csv")

#removing any edge sites 
meta_AB <- meta_AB %>%
  dplyr::filter(!grepl("Edge", UniqueFragID))

#setting correct structure: 
meta_AB$Forest_Group <- as.factor(meta_AB$Forest_Group)
meta_AB$Matrix_mean <- as.factor(meta_AB$Matrix_mean)

# Setting weight
vi_floor <- 0.2

#Make a tiny lookup of fragment -> sample size (use existing names)
frag_sizes <- meta_AB %>%
  select(UniqueFragID_site, Sample_size) %>%
  distinct()

#Join partner sample size (partner ID may be NA for pooled controls)
meta_AB <- meta_AB %>%
  left_join(frag_sizes,
            by = c("Partner_UniqueFragID" = "UniqueFragID_site"),
            suffix = c("", ".partner"))

#creating sample size columns for variance calculation
meta_AB <- meta_AB %>%
  mutate(
    nt_raw = Sample_size,                # focal (this dataset's site-level sample size)
    nc_raw = Sample_size.partner,        # partner sample size (joined)
    nt = as.numeric(nt_raw),
    nc = as.numeric(nc_raw)
  )

meta_AB <- meta_AB %>%
  mutate(
    nt_fix = ifelse(is.na(nt) | nt <= 0, 0.5, nt),
    nc_fix = ifelse(is.na(nc) | nc <= 0, 0.5, nc),
    vi_raw = 1 / nt_fix + 1 / nc_fix,
    vi = pmax(vi_raw, vi_floor),   # floor to prevent tiny variances
    se_obs = sqrt(vi)
  )

mod_data <- meta_AB %>% filter(is.finite(vi) & vi > 0)

#consistently orienting lnRR so positive means higher abundance in continuous forests 
meta_AB <- meta_AB %>%
  mutate(
    Comparison = if ("Comparison" %in% names(.)) Comparison else NA_character_,
    # safe normalisation of comparison names if needed
    Comparison = dplyr::case_when(
      !is.na(Comparison) ~ Comparison,
      !is.na(Treatment) & Treatment == "Frag" ~ "Frag_vs_Cont",
      TRUE ~ Comparison
    ),
    lnRR_for_plot = case_when(
      Comparison == "Frag_vs_Cont" ~ lnRR,     # frag / cont -> positive means fragment > continuous
      Comparison == "Cont_vs_Frag" ~ -lnRR,    # if you have inverse rows, flip
      Comparison == "Frag_vs_Frag" ~ -lnRR,    # flip so + means larger fragment higher (consistent)
      TRUE ~ lnRR                              # fallback (leave original)
    )
  )

#removing paper 34 and 52
meta_AB <-subset(meta_AB,Paper_ID!= 34)
meta_AB <-subset(meta_AB,Paper_ID!= 52)

```
### Separating out the data for the different comparison types, and ensuring the structure is correct for modeling:

```{r}
#selecting only frag vs continuous
meta_AB_FC <- meta_AB %>%
  # keep only Frag vs Cont contrasts
  filter(Comparison %in% c("Frag_vs_Cont", "Cont_vs_Frag")) %>%
  # orient so all effects mean "fragment relative to continuous"
  mutate(
    lnRR_dir = ifelse(Treatment == "Cont", -lnRR, lnRR)
  )

#inserting model:
#gaussian
model_AB_full  <- readRDS("m_abundance_CF_gaussian_new.rds")
summary(model_AB_full)

#student t
model_AB_full_student  <- readRDS("m_abundance_CF_student_new.rds")
summary(model_AB_full_student)

loo_ab_gaussian <- loo(model_AB_full)
loo_ab_student <- loo(model_AB_full_student)
loo_compare(loo_ab_gaussian, loo_ab_student)

```
# F-F comparisons: 

```{r}
meta_AB_FF <- meta_AB %>%
  filter(Comparison == "Frag_vs_Frag") %>%
  mutate(
    size_small = Fragment_size_ha,   # the smaller of the pair (from your construction)
    log_size_small = log(size_small)
  ) %>%
  filter(is.finite(log_size_small)) #creating a logged number that is the smallest fragment of the two being compared, because the size of the smaller fragment is the variable that changes between comparisons — the large fragment just acts as a reference.

#scaling log_size_small: 
meta_AB_FF <- meta_AB_FF %>%
  mutate(log_size_small_s = as.numeric(scale(log_size_small)))

model_AB_fragfrag_full  <- readRDS("m_abundance_FF_new.rds")
summary(model_AB_fragfrag_full)

```
### Now, we are going to prepare the full dataset, combining F-F and F-C comparisons, and assigning the value of 10K ha to continuous forest
```{r}
cf_area_ha <- 10000  # conventional area assigned to continuous forest

meta_AB_All <- meta_AB %>%
  mutate(
    .orig_comp = Comparison,
    
    # CHANGED: standardise comparison labels only.
    # No lnRR sign changes are made here.
    Comparison = case_when(
      .orig_comp %in% c("Frag_vs_Frag", "Frag_vs_Cont") ~ .orig_comp,
      .orig_comp == "Cont_vs_Frag" ~ "Frag_vs_Cont",
      TRUE ~ NA_character_
    ),
    
    # Make area columns numeric
    Fragment_size_ha = as.numeric(Fragment_size_ha),
    Partner_Fragment_size_ha = as.numeric(Partner_Fragment_size_ha),
    Cont_Fragment_size_ha = as.numeric(Cont_Fragment_size_ha)
  ) %>%
  
  # Keep only recognised contrast types
  filter(!is.na(Comparison)) %>%
  
  mutate(
    forest_area_ref_ha = case_when(
      Comparison == "Frag_vs_Frag" ~ pmax(Fragment_size_ha, Partner_Fragment_size_ha, na.rm = TRUE),
      Comparison == "Frag_vs_Cont" ~ cf_area_ha,
      TRUE ~ NA_real_
    ),
    
    # Replace Inf/-Inf from pmax when both values are missing
    forest_area_ref_ha = if_else(is.finite(forest_area_ref_ha), forest_area_ref_ha, NA_real_),
    
    # Log-transform and standardise for modelling
    log_forest_area_ref   = log(forest_area_ref_ha),
    log_forest_area_ref_s = as.numeric(scale(log_forest_area_ref))
  ) %>%
  
  # Drop rows that still lack a usable forest-area moderator
  filter(is.finite(log_forest_area_ref)) %>%
  
  # Remove helper column
  select(-.orig_comp)

  #scaling variables: 
meta_AB_All <- meta_AB_All %>%
  mutate(
    # numeric transforms
    Elevation_s   = as.numeric(scale(Elevation)),
    Rainfall_mm_s = as.numeric(scale(Rainfall_mm)),
    Latitude_s    = as.numeric(scale(Latitude)),
    Longitude_s   = as.numeric(scale(Longitude))
  )

#model without interaction term
meta_AB_all_full  <- readRDS("m_abundance_ALL_new.rds")
summary(meta_AB_all_full)

#model with interaction term (Comparison and the size of the reference forest area)
m_int_AB_ALL  <- readRDS("m_abundance_interaction_ALL_new.rds")
summary(m_int_AB_ALL)

#loo test 
loo_AB_no_interaction <- loo(meta_AB_all_full)
loo_AB_interaction <- loo(m_int_AB_ALL)
loo_compare(loo_AB_no_interaction, loo_AB_interaction)

mcmc_intervals(as_draws_df(m_int_AB_ALL), pars = vars(starts_with("b_")),
               prob = 0.95) +
  vline_at(0, linetype = "dashed")

#looking at the smallest fragment size as a moderator in the full dataset:

meta_AB_All <- meta_AB_All %>%
  mutate(
    log_size_small = case_when(
      Comparison == "Frag_vs_Frag" ~ log(pmin(Fragment_size_ha, Partner_Fragment_size_ha)),
      Comparison == "Frag_vs_Cont" ~ log(Fragment_size_ha),
      TRUE ~ NA_real_
    ),
    log_size_small_s = as.numeric(scale(log_size_small))
  )

m_int_AB_ALL_fragsize  <- readRDS("m_abundance_ALL_FragSize.rds")
summary(m_int_AB_ALL_fragsize)

```