### In this document, we will calculate the effect size for each of my datasets: species-level and assembly-level. We also look at vi (weights) and how many rows have missing sample sizes, and how many of those are being filled with the fallback value of 0.5.

```{r}
#Loading the necessary libraries
library(dplyr)
library(tidyr)
library(readr)
library(stringr)


#Loading the datasets:
species_data <- read_csv("species_level_data_cleaned.csv")
meta_SR <- read_csv("meta_SR_cleaned.csv")  
meta_AB <- read_csv("meta_AB_cleaned.csv")
meta_ACT <- read_csv("meta_ACT_cleaned.csv")  

colnames(species_data)
```

### Ensuring the structure of the datasets is correct and that the necessary columns for effect size calculation are present, and then ensuring eps is calculated for each dataset (starting wiht species-level):

```{r}
#Checking the structure of the datasets:
str(species_data)
str(meta_SR)
str(meta_AB)
str(meta_ACT)   #They all look ok.

# Species-level dataset: use Rate
species_data <- species_data %>%
  mutate(lnRR_input = Rate)

# Species Richness dataset: use raw counts
meta_SR <- meta_SR %>%
  mutate(lnRR_input = Response)

# Total Abundance dataset: use Rate
meta_AB <- meta_AB %>%
  mutate(lnRR_input = Rate)

# Activity dataset: use Rate
meta_ACT <- meta_ACT %>%
  mutate(lnRR_input = Rate)

#For species-level
eps <- min(species_data$lnRR_input[species_data$lnRR_input > 0], na.rm = TRUE) / 2

#sensitivity analysis: I wil then incorporate these into new lnrr caluclations.
eps1_2 <- eps / 2      # half the current epsilon
eps2   <- eps * 2      # double the current epsilon

```

### Equation for calculating the effect size for species-level data:

```{r}
#Creating the function to caluclate the effect size for species-level data using log response ratio (lnRR):
#species-specific
calculate_lnRR <- function(df, value_col = "Rate", eps = NULL) {
  library(dplyr)
  
  stopifnot(is.data.frame(df))
  if (!value_col %in% names(df)) stop("value_col not found in df")
  required <- c("Paper_ID","Species","Treatment","SiteUID","UniqueFragID","Fragment_size_ha")
  miss <- setdiff(required, names(df))
  if (length(miss)) stop("Missing columns: ", paste(miss, collapse = ", "))
  
  df <- as_tibble(df) %>% ungroup()
  
  if (is.null(eps)) {
    pos <- df[[value_col]][df[[value_col]] > 0]
    eps <- ifelse(length(pos) == 0, 1e-6, min(pos, na.rm = TRUE) / 2)
  }
  
  results <- list(); k <- 1L
  
  groups <- df %>% group_by(Paper_ID, Species) %>% group_split(.keep = TRUE)
  
  for (g in groups) {
    p_id <- unique(g$Paper_ID); sp <- unique(g$Species)
    if (length(p_id) != 1 || length(sp) != 1) next
    
    # ----- aggregate numeric stats per SiteUID -----
    agg_stats <- g %>%
      group_by(SiteUID) %>%
      summarise(
        Sample_size_sum = if ("Sample_size_sum" %in% names(.)) {
          sum(as.numeric(Sample_size_sum), na.rm = TRUE)
        } else if ("Sample_size" %in% names(.)) {
          sum(as.numeric(Sample_size), na.rm = TRUE)
        } else {
          NA_real_
        },
        Response = sum(as.numeric(Response), na.rm = TRUE),
        Mist_net_hours = sum(as.numeric(Mist_net_hours), na.rm = TRUE),
        .groups = "drop"
      )
    
    # ----- representative row per SiteUID to preserve metadata/traits -----
    site_repr <- g %>%
      group_by(SiteUID) %>%
      slice_head(n = 1) %>%  # choose first row as representative; change if you prefer another rule
      ungroup()
    
    # ----- join aggregated stats back onto representative rows -----
    site_tbl <- site_repr %>%
      # remove any pre-existing aggregated columns safely
      select(-any_of(c("Response","Mist_net_hours", "Sample_size_sum","site_value"))) %>%
      left_join(agg_stats, by = "SiteUID") %>%
      mutate(
        # preserve fragment and canonical columns from the representative row
        Fragment_size_ha = site_repr$Fragment_size_ha[match(SiteUID, site_repr$SiteUID)],
        Paper_ID = site_repr$Paper_ID[match(SiteUID, site_repr$SiteUID)],
        Species  = site_repr$Species[match(SiteUID, site_repr$SiteUID)],
        Treatment = site_repr$Treatment[match(SiteUID, site_repr$SiteUID)],
        UniqueFragID = site_repr$UniqueFragID[match(SiteUID, site_repr$SiteUID)]
      ) %>%
      # compute site_value using aggregated stats (fall back to mean of raw if no effort)
      mutate(
        site_value = ifelse(Mist_net_hours > 0,
                            Response / Mist_net_hours,
                            ifelse(value_col %in% names(site_repr),
                                   mean(site_repr[[value_col]], na.rm = TRUE),
                                   NA_real_))
      ) %>%
      distinct(SiteUID, .keep_all = TRUE)
    
    if (any(is.na(site_tbl$Treatment))) site_tbl$Treatment[is.na(site_tbl$Treatment)] <- "Unknown"
    
    frag_sites <- site_tbl %>% filter(Treatment == "Frag")
    cont_sites <- site_tbl %>% filter(Treatment == "Cont")
    
    ## Frag vs Cont
    if (nrow(frag_sites) > 0 && nrow(cont_sites) > 0) {
      total_effort <- sum(cont_sites$Mist_net_hours, na.rm = TRUE)
      pooled_cont_value <- if (is.finite(total_effort) && total_effort > 0) {
        sum(cont_sites$Response, na.rm = TRUE) / total_effort
      } else {
        # if cont_sites$site_value exists, use its mean; otherwise NA
        mean(cont_sites$site_value, na.rm = TRUE)
      }
      
      frag_keep <- frag_sites %>%
        filter(!((is.na(site_value) | site_value == 0) & (is.na(pooled_cont_value) | pooled_cont_value == 0)))
      
      if (nrow(frag_keep) > 0) {
        # pooled control sample-size info (if Sample_size_sum missing it becomes 0)
        pooled_cont_sample_sum <- sum(cont_sites$Sample_size_sum, na.rm = TRUE)
        pooled_cont_n_sites   <- nrow(cont_sites)
        lnrr_vals <- log((frag_keep$site_value + eps) / (pooled_cont_value + eps))
        
        out <- frag_keep %>%
          mutate(
            Cont_Paper_ID         = p_id,
            Cont_SiteUID          = paste0("Pooled_Cont__", p_id),
            Cont_UniqueFragID     = paste0("Pooled_Cont__", p_id),
            Cont_Fragment_size_ha = NA_real_,
            Cont_Value            = pooled_cont_value,
            Cont_Sample_size_sum  = pooled_cont_sample_sum,
            Cont_n_sites          = pooled_cont_n_sites,
            lnRR                  = lnrr_vals,
            Comparison            = "Frag_vs_Cont",
            Type                  = "Species_level"
          )
        
        # preserve original columns and add above new ones (no column removal)
        results[[k]] <- out
        k <- k + 1L
      }
    }
    
    ## Frag vs Frag
    if (nrow(frag_sites) > 1) {
      n <- nrow(frag_sites)
      pairs <- which(outer(1:n, 1:n, `<`), arr.ind = TRUE)
      if (nrow(pairs) > 0) {
        for (r in seq_len(nrow(pairs))) {
          i <- pairs[r,1]; j <- pairs[r,2]
          vi <- frag_sites$site_value[i]; vj <- frag_sites$site_value[j]
          if ((is.na(vi) || vi == 0) && (is.na(vj) || vj == 0)) next
          si <- frag_sites$Fragment_size_ha[i]; sj <- frag_sites$Fragment_size_ha[j]
          if (!is.na(si) && !is.na(sj) && si == sj) next
          
          if (is.na(si) || is.na(sj) || si < sj) {
            num <- vi; den <- vj
            base  <- frag_sites[i, , drop = FALSE]; other <- frag_sites[j, , drop = FALSE]
          } else {
            num <- vj; den <- vi
            base  <- frag_sites[j, , drop = FALSE]; other <- frag_sites[i, , drop = FALSE]
          }
          
          out <- base %>%
            mutate(
              Partner_Paper_ID         = other$Paper_ID,
              Partner_SiteUID          = other$SiteUID,
              Partner_UniqueFragID     = other$UniqueFragID,
              Partner_Fragment_size_ha = other$Fragment_size_ha,
              Partner_site_value       = other$site_value,
              Cont_Sample_size_sum     = other$Sample_size_sum,
              Cont_n_sites             = 1L,
              lnRR                     = log((num + eps) / (den + eps)),
              Comparison               = "Frag_vs_Frag",
              Type                     = "Species_level"
            )
          
          results[[k]] <- out
          k <- k + 1L
        }
      }
    }
  }
  
  if (k == 1L) return(tibble())
  bind_rows(results[1:(k - 1)])
}
``` 
### Altering the function so that we half and double the epsilon value to ensure that the effect size is not being driven by the choice of epsilon.

```{r}
#Testing the function with different epsilon values:
calculators <- list(
  calculate_lnrr_1_2 = function(df, value_col = "Rate") {
    calculate_lnRR(df, eps = eps1_2, value_col = value_col)
  },
  calculate_lnrr_orig = function(df, value_col = "Rate") {
    calculate_lnRR(df, eps = eps, value_col = value_col)
  },
  calculate_lnrr_2 = function(df, value_col = "Rate") {
    calculate_lnRR(df, eps = eps2, value_col = value_col)
  }
)
```

### Running the function on the species-level dataset for all three epsilon values and testing to see if it worked: 

```{r}
#Running for different epsilon values:
lnrr_orig <- calculate_lnRR(species_data, value_col = "Rate", eps = eps) #original epsilon
lnrr_double <- calculators$calculate_lnrr_2(species_data, value_col = "Rate") #double epsilon
lnrr_half <- calculators$calculate_lnrr_1_2(species_data, value_col = "Rate") #half epsilon

#Testing to see if it worked:
summary(lnrr_orig$lnRR - lnrr_half$lnRR)
mean(abs(lnrr_half$lnRR)) 
mean(abs(lnrr_orig$lnRR)) 
mean(abs(lnrr_double$lnRR)) 

#renaming original lnrr dataset to lnrr_correct for clarity:
lnrr_correct <- lnrr_orig
```

### Ensuring sample sizes are brought through correctly in the effect size calculation for species-level data, as the continuous forest sites were pooled.

```{r}
# ---------- frag_sizes build ----------
frag_sizes <- lnrr_correct %>%
  # start with an NA column to fill safely
  mutate(.ss_base = NA_real_) %>%
  { tmp <- .;
  # prefer Sample_size_sum if present, otherwise use Sample_size if present
  if ("Sample_size_sum" %in% names(tmp)) {
    tmp$.ss_base <- as.numeric(tmp$Sample_size_sum)
  } else if ("Sample_size" %in% names(tmp)) {
    tmp$.ss_base <- as.numeric(tmp$Sample_size)
  } else {
    tmp$.ss_base <- NA_real_
  }
  tmp
  } %>%
  # if a pooled control sample-size exists and the pooled control UID equals the row UID,
  # use the pooled control sample size (safe references)
  mutate(.ss_base = if (
    all(c("Cont_Sample_size_sum", "Cont_UniqueFragID") %in% names(.))
  ) {
    ifelse(!is.na(Cont_Sample_size_sum) & !is.na(Cont_UniqueFragID) & Cont_UniqueFragID == UniqueFragID,
           as.numeric(Cont_Sample_size_sum),
           .ss_base)
  } else .ss_base) %>%
  # aggregate: one row per UniqueFragID, take median of available sample-size proxies
  group_by(UniqueFragID) %>%
  summarise(
    Sample_size_sum = if (all(is.na(.ss_base))) NA_real_ else median(.ss_base, na.rm = TRUE),
    n_rows = n(),
    .groups = "drop"
  )

# If pooled controls were generated as Cont_UniqueFragID but not copied into UniqueFragID rows,
# create explicit pooled UniqueFragIDs from lnrr_correct pooled rows.
if (!("Pooled_Cont__" %in% substr(frag_sizes$UniqueFragID, 1, 13))) {
  pooled_rows <- lnrr_correct %>%
    filter(!is.na(Cont_UniqueFragID) & !is.na(Cont_Sample_size_sum)) %>%
    select(UniqueFragID = Cont_UniqueFragID, Sample_size_sum = Cont_Sample_size_sum) %>%
    distinct()
  if (nrow(pooled_rows) > 0) {
    frag_sizes <- bind_rows(frag_sizes, pooled_rows %>% mutate(n_rows = 1)) %>%
      group_by(UniqueFragID) %>%
      summarise(Sample_size_sum = median(Sample_size_sum, na.rm = TRUE), n_rows = sum(n_rows), .groups = "drop")
  }
}


# join partner sample size using Partner_UniqueFragID -> UniqueFragID
# produce Sample_size_sum.partner from the frag_sizes lookup
lnrr_with_partner_ss <- lnrr_correct %>%
  left_join(frag_sizes %>% rename(Sample_size_sum.partner = Sample_size_sum),
            by = c("Partner_UniqueFragID" = "UniqueFragID"))

df_name <- "lnrr_with_partner_ss"  
df <- get(df_name)
cat("Rows:", nrow(df), " Columns:", ncol(df), "\n\n")
print(names(df))

# Fill partner sample-size from pooled fields if join failed (i.e. for pooled controls that
#    were attached to the lnrr row rather than present as separate UniqueFragID rows)
# work on a copy to be safe 
lnrr_with_partner_ss <- df

# coerce/normalise the two target columns 
lnrr_with_partner_ss <- lnrr_with_partner_ss %>%
  dplyr::mutate(
    # focal sample-size:  Sample_size_sum
    Sample_size_sum = as.numeric(Sample_size_sum),
    
    # partner sample-size: use existing Sample_size_sum.partner if present,
    # otherwise fall back to Cont_Sample_size_sum 
    Sample_size_sum.partner = as.numeric(
      dplyr::coalesce(Sample_size_sum.partner, Cont_Sample_size_sum)
    )
  )
```

###     Computing variance (vi) for each lnRR

```{r}
#Compute variance (vi) for each lnRR using the formula: vi = (1/n1) + (1/n2)
vi_floor <- 0.2  # keep chosen floor

meta_species2 <- lnrr_with_partner_ss %>%
  mutate(
    nt_raw = Sample_size_sum,                 # focal fragment sample size sum (from lnrr base rows)
    nc_raw = Sample_size_sum.partner,         # partner fragment (joined or pooled)
    nt = as.numeric(nt_raw),
    nc = as.numeric(nc_raw),
    nt_fix = ifelse(is.na(nt) | nt <= 0, 0.5, nt),
    nc_fix = ifelse(is.na(nc) | nc <= 0, 0.5, nc),
    vi_raw = 1/nt_fix + 1/nc_fix,
    vi     = pmax(vi_raw, vi_floor),
    se_obs = sqrt(vi) 
  )

#Diagnostics: counts before/after and how many rows depend on fallback
diag <- list(
  n_rows_total = nrow(meta_species2),
  n_nc_missing = sum(is.na(meta_species2$nc)),
  n_nc_fallback = sum(meta_species2$nc_fix == 0.5, na.rm = TRUE),
  prop_nc_fallback = mean(meta_species2$nc_fix == 0.5, na.rm = TRUE),
  n_nt_missing = sum(is.na(meta_species2$nt)),
  n_nt_fallback = sum(meta_species2$nt_fix == 0.5, na.rm = TRUE)
)
cat("Rows total:", diag$n_rows_total, "\n")
cat("Rows with nc missing:", diag$n_nc_missing, "\n")
cat("Rows with nc_fix == 0.5 (fallback):", diag$n_nc_fallback, " (", round(100*diag$prop_nc_fallback,2), "%)\n", sep="")

# Compare the two vi definitions and show how many would change >50%
meta_species2 <- meta_species2 %>%
  mutate(
    vi_def1 = pmax(1/nt_fix + 1/nc_fix, vi_floor),   # current approach
    vi_def2 = pmax(1/(nt_fix + nc_fix), vi_floor),   # alternative
    rel_change = ifelse(is.finite(vi_def1) & vi_def1 > 0, (vi_def2 - vi_def1) / vi_def1, NA_real_)
  )

tbl_change <- table(abs(meta_species2$rel_change) > 0.5, useNA = "ifany")
cat("Rows with relative change > 50% between vi definitions:\n")
print(tbl_change)
cat("Percent > 50% change:", round(100 * sum(abs(meta_species2$rel_change) > 0.5, na.rm = TRUE) / nrow(meta_species2), 2), "%\n")

# Assign back into environment
lnrr_final <- meta_species2  
colnames(lnrr_final)  # check the new columns

n_distinct(lnrr_final$Species) #219 species represented!

#Writing csv: 
write.csv(lnrr_final, "effect_size_species.csv")
```

### Creating function to calculate effect size for assembly-level data (Species Richness)

```{r}
# ---- Define lnRR calculation function for SR----

calculate_lnRR_SR <- function(df, eps = NULL) {
  library(dplyr)
  
  df <- as_tibble(df) %>% ungroup()
  
  if (is.null(eps)) {
    pos <- df$Response[df$Response > 0] 
    eps <- ifelse(length(pos) == 0, 1e-6, min(pos, na.rm = TRUE) / 2) #add an epsilon value of half the smallest value on to values of zero (absence of species)
  }
  
  results <- list(); k <- 1L
  groups <- df %>% group_by(Paper_ID) %>% group_split(.keep = TRUE) #split data by study (Paper_ID)
  
  for (g in groups) {
    p_id <- unique(g$Paper_ID)
    if (length(p_id) != 1) next
    
    site_tbl <- g %>%
      group_by(SiteUID) %>%
      summarise(
        Paper_ID = first(Paper_ID),
        Treatment = first(Treatment),
        UniqueFragID = first(UniqueFragID),
        Fragment_size_ha = first(Fragment_size_ha),
        Response = sum(Response, na.rm = TRUE), # sum raw species counts across rows that share SiteUID
        Sample_size_sum = sum(Sample_size, na.rm = TRUE),
        Mist_net_hours = sum(Mist_net_hours, na.rm = TRUE), # for SR we use raw count as site_value (not rate)
        site_value = Response,
        .groups = "drop"
      )
    
    frag_sites <- site_tbl %>% filter(Treatment == "Frag")
    cont_sites <- site_tbl %>% filter(Treatment == "Cont")
    
    ## Frag vs Cont : tehre must be at least one frag value AND one cont value
    if (nrow(frag_sites) > 0 && nrow(cont_sites) > 0) {
      pooled_cont_value <- sum(cont_sites$Response, na.rm = TRUE)
      
      frag_keep <- frag_sites %>% #if there are two zero values, then drop that comparison
        filter(!((site_value == 0) & (pooled_cont_value == 0)))
      
      if (nrow(frag_keep) > 0) {
        out <- frag_keep %>%
          mutate(
            Cont_Paper_ID = p_id,
            Cont_SiteUID = paste0("Pooled_Cont__", p_id), #label for pooled continuous baseline
            Cont_UniqueFragID = NA_character_,
            Cont_Fragment_size_ha = NA_real_,
            Cont_Sample_size_sum = sum(cont_sites$Sample_size_sum),
            Cont_Value = pooled_cont_value,
            lnRR = log((site_value + eps) / (pooled_cont_value + eps)),
            Comparison = "Frag_vs_Cont",
            Type = "SR_level"
          )
        results[[k]] <- out; k <- k + 1L
      }
    }
    
    ## Frag vs Frag : 
    # create all unique unordered pairs (i < j) of fragment sites within the same paper
    if (nrow(frag_sites) > 1) {
      n <- nrow(frag_sites)
      pairs <- which(outer(1:n, 1:n, `<`), arr.ind = TRUE)
      
      for (r in seq_len(nrow(pairs))) {
        i <- pairs[r,1]; j <- pairs[r,2]
        vi <- frag_sites$site_value[i]; vj <- frag_sites$site_value[j]
        if (vi == 0 & vj == 0) next # skip pairs where both values are zero (0/0 undefined)
        
        si <- frag_sites$Fragment_size_ha[i]; sj <- frag_sites$Fragment_size_ha[j]
        if (!is.na(si) && !is.na(sj) && si == sj) next ## skip pairs with identical sizes (ambiguous "smaller vs larger" ordering)
        
        ## decide orientation so that "num" is the smaller fragment and "den" is the larger
        # if either size is NA we default to the first index (i) being treated as 'smaller'
        
        if (is.na(si) || is.na(sj) || si < sj) {
          num <- vi; den <- vj
          base <- frag_sites[i,]; other <- frag_sites[j,]
        } else {
          num <- vj; den <- vi
          base <- frag_sites[j,]; other <- frag_sites[i,]
        }
        # build output row: base fragment augmented with partner info and lnRR
        out <- base %>%
          mutate(
            Partner_Paper_ID = p_id,
            Partner_SiteUID = other$SiteUID,
            Partner_UniqueFragID = other$UniqueFragID,
            Partner_Fragment_size_ha = other$Fragment_size_ha,
            Partner_Sample_size_sum = other$Sample_size_sum,
            Partner_site_value = other$site_value,
            lnRR = log((num + eps) / (den + eps)),
            Comparison = "Frag_vs_Frag",
            Type = "SR_level"
          )
        
        results[[k]] <- out; k <- k + 1L
      }
    }
  }
  
  if (k == 1L) return(tibble())
  bind_rows(results[1:(k - 1)])
}

# SR epsilon (raw species richness counts)
eps <- min(meta_SR$Response[meta_SR$Response > 0], na.rm = TRUE) / 2

eps1_2 <- eps / 2
eps2 <- eps * 2

# ---- SR ----
calculators <- list(
  calculate_lnrr_1_2 = function(df) {
    calculate_lnRR_SR(df, eps = eps1_2)
  },
  calculate_lnrr_orig = function(df) {
    calculate_lnRR_SR(df, eps = eps)
  },
  calculate_lnrr_2 = function(df) {
    calculate_lnRR_SR(df, eps = eps2)
  }
)

meta_SR <- meta_SR %>% mutate(Value = Response)

```

### Applying the function to the SR dataset and checking the results and joining back to the original dataframe to retain columns:

```{r}
#Calculating lnRR for SR dataset with different epsilon values:
lnRR_SR  <- calculate_lnRR_SR(meta_SR)
lnRR_SR_half <- calculators$calculate_lnrr_1_2(meta_SR) #half epsilon
lnRR_SR_dbl <- calculators$calculate_lnrr_2(meta_SR) #double epsilon

# Joining back to the original dataframe to retain columns:
site_lookup_SR <- meta_SR %>%
  distinct(Paper_ID, SiteUID, .keep_all = TRUE)

#Original epsilon:
lnRR_SR_full <- lnRR_SR %>%
  left_join(site_lookup_SR,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))

#Half epsilon:
lnRR_SR_half <- lnRR_SR_half %>%
  left_join(site_lookup_SR,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))

#Double epsilon:
lnRR_SR_dbl <- lnRR_SR_dbl %>%
  left_join(site_lookup_SR,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))

 #Writing SR csvs: 
 write.csv(lnRR_SR_full, "effect_size_totalSR.csv")   #Original epsilon
 write.csv(lnRR_SR_half, "effect_size_totalSR_half_eps.csv") #Half epsilon
 write.csv(lnRR_SR_dbl, "effect_size_totalSR_double_eps.csv") #Double epsilon        

```

### Doing the same for the AB dataset:

```{r}
# Creating the lnrr function for AB dataset:
calculate_lnRR_AB <- function(df, eps = NULL) {
  library(dplyr)
  
  required <- c("Paper_ID","Treatment","SiteUID","UniqueFragID","Fragment_size_ha",
                "Response","Mist_net_hours","Rate")
  miss <- setdiff(required, names(df))
  if (length(miss)) stop("Missing columns: ", paste(miss, collapse = ", "))
  
  df <- as_tibble(df) %>% ungroup()
  
  if (is.null(eps)) {
    pos <- df$Rate[df$Rate > 0]
    eps <- ifelse(length(pos) == 0, 1e-6, min(pos, na.rm = TRUE) / 2)
  }
  
  results <- list(); k <- 1L
  groups <- df %>% group_by(Paper_ID) %>% group_split(.keep = TRUE)
  
  for (g in groups) {
    p_id <- unique(g$Paper_ID)
    if (length(p_id) != 1) next
    
    site_tbl <- g %>%
      group_by(SiteUID) %>%
      summarise(
        Paper_ID = first(Paper_ID),
        Treatment = first(Treatment),
        UniqueFragID = first(UniqueFragID),
        Fragment_size_ha = first(Fragment_size_ha),
        Response = sum(Response, na.rm = TRUE),
        Sample_size_sum = sum(Sample_size, na.rm = TRUE),
        Mist_net_hours = sum(Mist_net_hours, na.rm = TRUE),
        site_value = ifelse(Mist_net_hours > 0,
                            Response / Mist_net_hours,
                            mean(Rate, na.rm = TRUE)),
        .groups = "drop"
      )
    
    frag_sites <- site_tbl %>% filter(Treatment == "Frag")
    cont_sites <- site_tbl %>% filter(Treatment == "Cont")
    
    ## Frag vs Cont
    if (nrow(frag_sites) > 0 && nrow(cont_sites) > 0) {
      total_effort <- sum(cont_sites$Mist_net_hours, na.rm = TRUE)
      pooled_cont_value <- if (total_effort > 0) {
        sum(cont_sites$Response, na.rm = TRUE) / total_effort
      } else {
        mean(cont_sites$site_value, na.rm = TRUE)
      }
      
      frag_keep <- frag_sites %>%
        filter(!((site_value == 0) & (pooled_cont_value == 0)))
      
      if (nrow(frag_keep) > 0) {
        out <- frag_keep %>%
          mutate(
            Cont_Paper_ID = p_id,
            Cont_SiteUID = paste0("Pooled_Cont__", p_id),
            Cont_UniqueFragID = NA_character_,
            Cont_Fragment_size_ha = NA_real_,
            Cont_Sample_size_sum = sum(cont_sites$Sample_size_sum),
            Cont_Value = pooled_cont_value,
            lnRR = log((site_value + eps) / (pooled_cont_value + eps)),
            Comparison = "Frag_vs_Cont",
            Type = "AB_level"
          )
        results[[k]] <- out; k <- k + 1L
      }
    }
    
    ## Frag vs Frag
    if (nrow(frag_sites) > 1) {
      n <- nrow(frag_sites)
      pairs <- which(outer(1:n, 1:n, `<`), arr.ind = TRUE)
      
      for (r in seq_len(nrow(pairs))) {
        i <- pairs[r,1]; j <- pairs[r,2]
        vi <- frag_sites$site_value[i]
        vj <- frag_sites$site_value[j]
        
        if ( (is.na(vi) || vi == 0) && (is.na(vj) || vj == 0) ) next
        
        si <- frag_sites$Fragment_size_ha[i]; sj <- frag_sites$Fragment_size_ha[j]
        if (!is.na(si) && !is.na(sj) && si == sj) next
        
        if (is.na(si) || is.na(sj) || si < sj) {
          num <- vi; den <- vj
          base <- frag_sites[i,]; other <- frag_sites[j,]
        } else {
          num <- vj; den <- vi
          base <- frag_sites[j,]; other <- frag_sites[i,]
        }
        
        out <- base %>%
          mutate(
            Partner_Paper_ID = p_id,
            Partner_SiteUID = other$SiteUID,
            Partner_UniqueFragID = other$UniqueFragID,
            Partner_Fragment_size_ha = other$Fragment_size_ha,
            Partner_site_value = other$site_value,
            Partner_Sample_size_sum = other$Sample_size_sum,
            lnRR = log((num + eps) / (den + eps)),
            Comparison = "Frag_vs_Frag",
            Type = "AB_level"
          )
        
        results[[k]] <- out; k <- k + 1L
      }
    }
  }
  
  if (k == 1L) return(tibble())
  bind_rows(results[1:(k - 1)])
}

# AB epsilon (capture rates)
eps <- min(meta_AB$Rate[meta_AB$Rate > 0], na.rm = TRUE) / 2

eps1_2 <- eps / 2
eps2 <- eps * 2

# ---- AB ----
calculators_AB <- list(
  calculate_lnrr_1_2 = function(df) {
    calculate_lnRR_AB(df, eps = eps1_2)
  },
  calculate_lnrr_orig = function(df) {
    calculate_lnRR_AB(df, eps = eps)
  },
  calculate_lnrr_2 = function(df) {
    calculate_lnRR_AB(df, eps = eps2)
  }
)
```

### Applying function to AB dataset and checking results:

```{r}
meta_AB <- meta_AB %>% mutate(Value = Rate)
lnRR_AB <- calculate_lnRR_AB(meta_AB)

# Applying different epsilon values:
lnRR_AB_half <- calculators_AB$calculate_lnrr_1_2(meta_AB) #half epsilon
lnRR_AB_dbl <- calculators_AB$calculate_lnrr_2(meta_AB) #double epsilon

#joining back to original to retain columns: 
site_lookup_AB <- meta_AB %>%
  distinct(Paper_ID, SiteUID, .keep_all = TRUE)

#Original epsilon:
lnRR_AB_full <- lnRR_AB %>%
  left_join(site_lookup_AB,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))    
#Half epsilon:
lnRR_AB_half <- lnRR_AB_half %>%
  left_join(site_lookup_AB,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))
#Double epsilon:
lnRR_AB_dbl <- lnRR_AB_dbl %>%
  left_join(site_lookup_AB,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))
```

### Writing AB csvs:

```{r}
write.csv(lnRR_AB_full, "effect_size_totalAB.csv")   #Original epsilon
write.csv(lnRR_AB_half, "effect_size_totalAB_half_eps.csv") #Half epsilon
write.csv(lnRR_AB_dbl, "effect_size_totalAB_double_eps.csv") #Double epsilon
```

### And lastly doing the same for the ACT dataset:

```{r}
#Function for activity: 

calculate_lnRR_activity <- function(df,
                                    value_col = "Rate",
                                    activity_count_col = "Activity",
                                    effort_col = "Mist_net_hours",
                                    eps = NULL,
                                    eps_floor = 0.01) {
  
  if (!requireNamespace("dplyr", quietly = TRUE)) stop("dplyr required")
  if (!requireNamespace("tibble", quietly = TRUE)) stop("tibble required")
  library(dplyr)
  library(tibble)
  
  stopifnot(is.data.frame(df))
  if (!value_col %in% names(df) && !activity_count_col %in% names(df)) {
    stop("Neither value_col nor activity_count_col found in df")
  }
  
  required <- c("Paper_ID", "Species", "Treatment", "SiteUID", "UniqueFragID", "Fragment_size_ha")
  miss <- setdiff(required, names(df))
  if (length(miss)) stop("Missing columns: ", paste(miss, collapse = ", "))
  
  # safe accessor for single-row 'other' data.frames used in mutate
  safe_col <- function(df_row, col_name) {
    # return NA_real_ for missing column, else column value (preserve type if present)
    if ((is.data.frame(df_row) || tibble::is_tibble(df_row)) && col_name %in% names(df_row)) {
      return(df_row[[col_name]])
    }
    return(NA_real_)
  }
  
  df <- as_tibble(df) %>% ungroup()
  
  # compute eps (continuity correction) if NULL using dataset-wide positive rates or fallback floor
  if (is.null(eps)) {
    pos_vals <- numeric(0)
    if (value_col %in% names(df)) {
      vals <- df[[value_col]]
      pos_vals <- c(pos_vals, vals[is.finite(vals) & vals > 0])
    }
    if (activity_count_col %in% names(df) && !is.null(effort_col) && effort_col %in% names(df)) {
      rates_from_counts <- with(df, ifelse(!is.na(.data[[activity_count_col]]) & .data[[effort_col]] > 0,
                                           .data[[activity_count_col]] / .data[[effort_col]], NA_real_))
      pos_vals <- c(pos_vals, rates_from_counts[is.finite(rates_from_counts) & rates_from_counts > 0])
    }
    if (length(pos_vals) == 0) {
      eps <- eps_floor
      message("No positive site rates found in df; using eps = eps_floor = ", eps)
    } else {
      min_pos <- min(pos_vals, na.rm = TRUE)
      eps <- max(min_pos / 2, eps_floor)
      message("Found min positive rate = ", format(min_pos, digits = 6),
              "; using eps = ", format(eps, digits = 6))
    }
  }
  
  results <- list(); k <- 1L
  groups <- df %>% group_by(Paper_ID, Species) %>% group_split(.keep = TRUE)
  
  for (g in groups) {
    p_id <- unique(g$Paper_ID); sp <- unique(g$Species)
    if (length(p_id) != 1 || length(sp) != 1) next
    
    # 1) aggregate per SiteUID: sample sums / counts / effort
    agg_stats <- g %>%
      group_by(SiteUID) %>%
      summarise(
        Sample_size_sum = if ("Sample_size_sum" %in% names(.)) sum(as.numeric(Sample_size_sum), na.rm = TRUE)
        else if ("Sample_size" %in% names(.)) sum(as.numeric(Sample_size), na.rm = TRUE)
        else NA_real_,
        Activity_sum = if (activity_count_col %in% names(.)) sum(as.numeric(.data[[activity_count_col]]), na.rm = TRUE) else NA_real_,
        Effort_sum = if (!is.null(effort_col) && effort_col %in% names(.)) sum(as.numeric(.data[[effort_col]]), na.rm = TRUE) else NA_real_,
        .groups = "drop"
      )
    
    # compute site_value per SiteUID (prefer provided value_col)
    if (value_col %in% names(g)) {
      site_values <- g %>%
        group_by(SiteUID) %>%
        summarise(site_value = mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
    } else {
      site_values <- agg_stats %>%
        mutate(site_value = ifelse(!is.na(Activity_sum) & !is.na(Effort_sum) & Effort_sum > 0,
                                   Activity_sum / Effort_sum, NA_real_)) %>%
        select(SiteUID, site_value)
    }
    
    # representative row to preserve metadata
    site_repr <- g %>%
      group_by(SiteUID) %>%
      slice_head(n = 1) %>%
      ungroup()
    
    # join aggregated stats and site_values onto representative rows
    site_tbl <- site_repr %>%
      left_join(agg_stats, by = "SiteUID") %>%
      left_join(site_values, by = "SiteUID") %>%
      mutate(
        Fragment_size_ha = site_repr$Fragment_size_ha[match(SiteUID, site_repr$SiteUID)],
        Paper_ID = site_repr$Paper_ID[match(SiteUID, site_repr$SiteUID)],
        Species  = site_repr$Species[match(SiteUID, site_repr$SiteUID)],
        Treatment = site_repr$Treatment[match(SiteUID, site_repr$SiteUID)],
        UniqueFragID = site_repr$UniqueFragID[match(SiteUID, site_repr$SiteUID)]
      ) %>%
      distinct(SiteUID, .keep_all = TRUE)
    
    if (any(is.na(site_tbl$Treatment))) site_tbl$Treatment[is.na(site_tbl$Treatment)] <- "Unknown"
    
    frag_sites <- site_tbl %>% filter(Treatment == "Frag")
    cont_sites <- site_tbl %>% filter(Treatment == "Cont")
    
    ## Frag vs Cont: compute pooled CONT value sensibly (prefer count/effort pooling)
    if (nrow(frag_sites) > 0 && nrow(cont_sites) > 0) {
      pooled_cont_value <- NA_real_
      pooled_cont_count  <- sum(cont_sites$Activity_sum, na.rm = TRUE)
      pooled_cont_effort <- sum(cont_sites$Effort_sum, na.rm = TRUE)
      
      if (!is.na(pooled_cont_effort) && pooled_cont_effort > 0 && !all(is.na(cont_sites$Activity_sum))) {
        # use count-based pooling; continuity correction if pooled count == 0
        if (!is.na(pooled_cont_count) && pooled_cont_count == 0) {
          pooled_cont_value <- (pooled_cont_count + 0.5) / pooled_cont_effort
          pooled_cont_zero_correction <- TRUE
        } else {
          pooled_cont_value <- pooled_cont_count / pooled_cont_effort
          pooled_cont_zero_correction <- FALSE
        }
      } else if (value_col %in% names(cont_sites) && !all(is.na(cont_sites[[value_col]]))) {
        if (!all(is.na(cont_sites$Effort_sum)) && sum(cont_sites$Effort_sum, na.rm = TRUE) > 0) {
          pooled_cont_value <- weighted.mean(cont_sites[[value_col]],
                                             w = ifelse(is.na(cont_sites$Effort_sum), 0, cont_sites$Effort_sum),
                                             na.rm = TRUE)
        } else {
          pooled_cont_value <- mean(cont_sites[[value_col]], na.rm = TRUE)
        }
        pooled_cont_zero_correction <- FALSE
      } else if (!all(is.na(cont_sites$site_value))) {
        pooled_cont_value <- mean(cont_sites$site_value, na.rm = TRUE)
        pooled_cont_zero_correction <- FALSE
      } else {
        pooled_cont_value <- NA_real_
        pooled_cont_zero_correction <- NA
      }
      
      if (!is.na(pooled_cont_value)) {
        frag_keep <- frag_sites %>%
          filter(!((is.na(site_value) | site_value == 0) & (is.na(pooled_cont_value) | pooled_cont_value == 0)))
        
        if (nrow(frag_keep) > 0) {
          pooled_cont_sample_sum <- sum(cont_sites$Sample_size_sum, na.rm = TRUE)
          pooled_cont_n_sites   <- nrow(cont_sites)
          lnrr_vals <- log((frag_keep$site_value + eps) / (pooled_cont_value + eps))
          
          out <- frag_keep %>%
            mutate(
              Cont_Paper_ID         = p_id,
              Cont_SiteUID          = paste0("Pooled_Cont__", p_id),
              Cont_UniqueFragID     = paste0("Pooled_Cont__", p_id),
              Cont_Fragment_size_ha = NA_real_,
              Cont_Value            = pooled_cont_value,
              Cont_Sample_size_sum  = pooled_cont_sample_sum,
              Cont_n_sites          = pooled_cont_n_sites,
              Cont_zero_correction  = ifelse(is.na(pooled_cont_zero_correction), FALSE, pooled_cont_zero_correction),
              lnRR                  = lnrr_vals,
              Comparison            = "Frag_vs_Cont",
              Type                  = "Activity_level"
            )
          
          results[[k]] <- out
          k <- k + 1L
        }
      }
    }
    
    ## Frag vs Frag: pairwise comparisons among fragments (respecting fragment size ordering)
    if (nrow(frag_sites) > 1) {
      n <- nrow(frag_sites)
      pairs <- which(outer(1:n, 1:n, `<`), arr.ind = TRUE)
      if (nrow(pairs) > 0) {
        for (r in seq_len(nrow(pairs))) {
          i <- pairs[r,1]; j <- pairs[r,2]
          vi <- frag_sites$site_value[i]; vj <- frag_sites$site_value[j]
          if ((is.na(vi) || vi == 0) && (is.na(vj) || vj == 0)) next
          si <- frag_sites$Fragment_size_ha[i]; sj <- frag_sites$Fragment_size_ha[j]
          if (!is.na(si) && !is.na(sj) && si == sj) next
          
          if (is.na(si) || is.na(sj) || si < sj) {
            num <- vi; den <- vj
            base  <- frag_sites[i, , drop = FALSE]; other <- frag_sites[j, , drop = FALSE]
          } else {
            num <- vj; den <- vi
            base  <- frag_sites[j, , drop = FALSE]; other <- frag_sites[i, , drop = FALSE]
          }
          
          out <- base %>%
            mutate(
              Partner_Paper_ID         = safe_col(other, "Paper_ID"),
              Partner_SiteUID          = safe_col(other, "SiteUID"),
              Partner_UniqueFragID     = safe_col(other, "UniqueFragID"),
              Partner_Fragment_size_ha = safe_col(other, "Fragment_size_ha"),
              Partner_site_value       = safe_col(other, "site_value"),
              Cont_Sample_size_sum     = dplyr::coalesce(safe_col(other, "Cont_Sample_size_sum"),
                                                         safe_col(other, "Sample_size_sum")),
              Cont_n_sites             = 1L,
              lnRR                     = log((num + eps) / (den + eps)),
              Comparison               = "Frag_vs_Frag",
              Type                     = "Activity_level"
            )
          
          results[[k]] <- out
          k <- k + 1L
        }
      }
    }
  } # end loop groups
  
  if (k == 1L) return(tibble::tibble())
  dplyr::bind_rows(results[1:(k - 1)])
}

n_distinct(meta_ACT$Paper_ID) #13

# ACT epsilon (activity rates with conservative floor)
eps <- max(
  min(meta_ACT$Rate[meta_ACT$Rate > 0], na.rm = TRUE) / 2,
  0.01
)

eps1_2 <- eps / 2
eps2 <- eps * 2

#Create calculators list for ACT dataset:
calculators_ACT <- list(
  calculate_lnrr_1_2 = function(df) {
    calculate_lnRR_activity(df, value_col = "Rate", eps = eps1_2)
  },
  calculate_lnrr_orig = function(df) {
    calculate_lnRR_activity(df, value_col = "Rate", eps = eps)
  },
  calculate_lnrr_2 = function(df) {
    calculate_lnRR_activity(df, value_col = "Rate", eps = eps2)
  }
)

```

### Applying function to ACT dataset and checking results:

```{r}
meta_ACT <- meta_ACT %>% mutate(Value = Rate)
lnRR_ACT <- calculate_lnRR_activity(meta_ACT)

# Applying different epsilon values:
lnRR_ACT_half <- calculators_ACT$calculate_lnrr_1_2(meta_ACT) #half epsilon
lnRR_ACT_dbl <- calculators_ACT$calculate_lnrr_2(meta_ACT) #double epsilon

#joining back to original to retain columns:
site_lookup_ACT <- meta_ACT %>%
  distinct(Paper_ID, SiteUID, .keep_all = TRUE) 

#Original epsilon:
lnRR_ACT_full <- lnRR_ACT %>%
  left_join(site_lookup_ACT,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))

#Half epsilon:
lnRR_ACT_half <- lnRR_ACT_half %>%
  left_join(site_lookup_ACT,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))

#Double epsilon:
lnRR_ACT_dbl <- lnRR_ACT_dbl %>%
  left_join(site_lookup_ACT,
            by = c("Paper_ID", "SiteUID"),
            suffix = c("", "_site"))    
```
### Writing ACT csvs:

```{r}
write.csv(lnRR_ACT_full, "effect_size_totalactivity.csv")   #Original epsilon
write.csv(lnRR_ACT_half, "effect_size_totalactivity_half_eps.csv") #Half epsilon
write.csv(lnRR_ACT_dbl, "effect_size_totalactivity_double_eps.csv") #Double epsilon
```