#load packages: 

```{r}
libs <- c(
  "brms", "projpred", "tidybayes", "bayesplot", "emmeans",
  "posterior", "performance", "loo", "car", "dplyr",
  "ggplot2", "tidyr", "gt"
)
lapply(libs, library, character.only = TRUE)
```

### Species-level sensitivity analysis: Here I will be testing whether halving or doubleing the value of epsilon (the small constant added to the sampling variance to prevent zero variances) has a significant effect on the posterior estimates for one of the main predictors (body mass) within my species level trait model.

```{r}
#C-F contrasts

#I want to compare the two models: 
m_SL_gaussian_FC_single  <- readRDS("m_batTraits_species.gaussian_CF.rds")
m_SL_gaussian_FC_hlfeps  <- readRDS("m_batTraits_species.gaussian_CF_halfeps.rds")
m_SL_gaussian_FC_dbleps  <- readRDS("m_batTraits_species.gaussian_CF_doubleeps.rds")

#look at posterior draws: 
draws_orig <- as_draws_df(m_SL_gaussian_FC_single)
draws_half <- as_draws_df(m_SL_gaussian_FC_hlfeps)
draws_double <- as_draws_df(m_SL_gaussian_FC_dbleps)

variables <- names(as_draws_df(m_SL_gaussian_FC_single))
variables[grepl("^b_", variables)]

params <- c(
  "b_Intercept",
  "b_Averagemass_g_s",
  "b_Conservation_statusDD",
  "b_Conservation_statusEN",
  "b_Conservation_statusNT",
  "b_Conservation_statusVU",
  "b_Habitat_affinity_newBackgroundMclutterMadapted",
  "b_Habitat_affinity_newEdgeMadapted",
  "b_Habitat_affinity_newOpenMadapted",
  "b_Habitat_affinity_newWaterMadapted",
  "b_DietCarnivore",
  "b_DietFrugivore",
  "b_DietInsectivore",
  "b_DietOmnivore",
  "b_DietPiscivore",
  "b_DietSanguivore"
)

post_orig <- draws_orig %>% select(all_of(params))
post_half <- draws_half %>% select(all_of(params))
post_double <- draws_double %>% select(all_of(params))

#overlaying density plots: 

plot_df <- bind_rows(
  post_orig %>%
    pivot_longer(cols = everything(),
                 names_to = "Parameter",
                 values_to = "Value") %>%
    mutate(Model = "Original"),
  post_half %>%
    pivot_longer(cols = everything(),
                 names_to = "Parameter",
                 values_to = "Value") %>%
    mutate(Model = "Half_epsilon"), 
   post_double %>%
     pivot_longer(cols = everything(),
                  names_to = "Parameter",
                  values_to = "Value") %>%
     mutate(Model = "Double_epsilon"),
)

#body mass plot: 
p_mass <- ggplot(
  plot_df %>% filter(Parameter == "b_Averagemass_g_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Body Mass effect)",
    y = "Density"
  ) +
  theme_minimal()
p_mass

#making a summary table: 
summ_fun <- function(df) {
  df %>%
    summarise(across(everything(),
                     list(
                       mean  = mean,
                       lower = ~quantile(.x, 0.025),
                       upper = ~quantile(.x, 0.975)
                     )
    ))
}

#Creating a summary table
sum_orig <- summ_fun(post_orig)
sum_half <- summ_fun(post_half)
sum_double <- summ_fun(post_double)


#reshape: 
make_table <- function(sum_df, model_name) {
  tibble(
    Parameter = rep(params, each = 1),
    Mean  = as.numeric(sum_df[1, seq(1, length(params)*3, 3)]),
    Lower = as.numeric(sum_df[1, seq(2, length(params)*3, 3)]),
    Upper = as.numeric(sum_df[1, seq(3, length(params)*3, 3)]),
    Model = model_name
  )
}

tab_orig <- make_table(sum_orig, "Original")
tab_half <- make_table(sum_half, "Half_epsilon")
tab_double <- make_table(sum_double, "Double_epsilon")


summary_table <- bind_rows(tab_orig, tab_half, tab_double) #, tab_double

#compute difference in posterior means: 
mean_diff <- colMeans(post_half) - colMeans(post_orig)

summary_table <- summary_table %>%
  mutate(
    Delta_mean = Mean - Mean[Model == "Original"]
  )

summary_table_wide_SLeps <- summary_table %>%
  mutate(
    Estimate = sprintf("%.2f [%.2f, %.2f]", Mean, Lower, Upper)
  ) %>%
  select(Parameter, Model, Estimate) %>%
  pivot_wider(
    names_from = Model,
    values_from = Estimate
  )

summary_table_wide_SLeps %>%
  mutate(
    Parameter = recode(
      Parameter,
      "b_Intercept" = "Intercept",
      "b_Averagemass_g_s" = "Body mass",
      "b_Conservation_statusDD" = "Conservation status: DD",
      "b_Conservation_statusEN" = "Conservation status: EN",
      "b_Conservation_statusNT" = "Conservation status: NT",
      "b_Conservation_statusVU" = "Conservation status: VU",
      "b_Habitat_affinity_newBackgroundMclutterMadapted" = "Habitat affinity: background clutter-adapted",
      "b_Habitat_affinity_newEdgeMadapted" = "Habitat affinity: edge-adapted",
      "b_Habitat_affinity_newOpenMadapted" = "Habitat affinity: open-adapted",
      "b_Habitat_affinity_newWaterMadapted" = "Habitat affinity: water-adapted",
      "b_DietCarnivore" = "Diet: carnivore",
      "b_DietFrugivore" = "Diet: frugivore",
      "b_DietInsectivore" = "Diet: insectivore",
      "b_DietOmnivore" = "Diet: omnivore",
      "b_DietPiscivore" = "Diet: piscivore",
      "b_DietSanguivore" = "Diet: sanguivore"
    )
  ) %>%
  gt(rowname_col = "Parameter") %>%
  tab_header(
    title = "Sensitivity analysis: continuity-correction epsilon"
  ) %>%
  cols_label(
    Original = "Original epsilon",
    Half_epsilon = "Half epsilon",
    Double_epsilon = "Double epsilon"
  ) %>%
  tab_options(
    table.font.size = px(13),
    row.striping.include_table_body = TRUE
  )
```

### Sensitivity analysis on Species Richness models: Here we will be testing whether halving or doubleing the value of epsilon (the small constant added to the sampling variance to prevent zero variances) has a significant effect on the posterior estimates for one of the main predictors (elevation) within my species richness model.
```{r}
#C-F contrasts

m_SR_gaussian_FC_single  <- readRDS("model_sr_full_gaussian.rds")
m_SR_gaussian_FC_dbleps  <- readRDS("model_sr_full_gaussian_dbleps.rds")
m_SR_gaussian_FC_hlfeps  <- readRDS("model_sr_full_gaussian_hlfeps.rds")

variables <- names(as_draws_df(m_SR_gaussian_FC_single))
variables[grepl("^b_", variables)]

#look at posterior draws: 
draws_single <- as_draws_df(m_SR_gaussian_FC_single)
draws_half <- as_draws_df(m_SR_gaussian_FC_hlfeps)
draws_double <- as_draws_df(m_SR_gaussian_FC_dbleps)

# parameters to compare
params <- c(
  "b_Intercept",
  "b_Elevation_s",
  "b_Rainfall_mm_s",
  "b_Latitude_s",
  "b_Longitude_s",
  "b_Matrix_meanOpen_land",
  "b_Matrix_meanProduction_forests",
  "b_Matrix_meanSecondaryforest",
  "b_Matrix_meanWater",
  "b_Forest_GroupTropical_Lowland",
  "b_Forest_GroupTropical_Montane"
)


#Summarise all fixed effects across epsilon choices and make a table:
summ_fun <- function(draws, model_name) {
  draws %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    group_by(Parameter) %>%
    summarise(
      Mean  = mean(Value),
      Lower = quantile(Value, 0.025),
      Upper = quantile(Value, 0.975),
      .groups = "drop"
    ) %>%
    mutate(Model = model_name)
}

# create summary tables
tab_single <- summ_fun(draws_single, "Original")
tab_half   <- summ_fun(draws_half, "Half_epsilon")
tab_double <- summ_fun(draws_double, "Double_epsilon")

# combine long-format tables
summary_table <- bind_rows(
  tab_single,
  tab_half,
  tab_double
)

# pivot wider for readability
summary_table_wide_SReps <- summary_table %>%
  
  mutate(
    Estimate = sprintf("%.2f [%.2f, %.2f]",
                       Mean, Lower, Upper)
  ) %>%
  
  select(Parameter, Model, Estimate) %>%
  
  pivot_wider(
    names_from = Model,
    values_from = Estimate
  )


summary_table_wide_SReps %>%
  
  mutate(
    Parameter = recode(
      Parameter,
      "b_Intercept" = "Intercept",
      "b_Elevation_s" = "Elevation",
      "b_Rainfall_mm_s" = "Rainfall",
      "b_Latitude_s" = "Latitude",
      "b_Longitude_s" = "Longitude",
      "b_Matrix_meanOpen_land" = "Open land",
      "b_Matrix_meanProduction_forests" = "Production forests",
      "b_Matrix_meanSecondaryforest" = "Secondary forest",
      "b_Matrix_meanWater" = "Water",
      "b_Forest_GroupTropical_Lowland" = "Tropical lowland",
      "b_Forest_GroupTropical_Montane" = "Tropical montane"
    )
  ) %>%
  
  gt(rowname_col = "Parameter") %>%
  
  tab_header(
    title = "Sensitivity analysis: epsilon correction"
  ) %>%
  
  cols_label(
    Original = "Original epsilon",
    Half_epsilon = "Half epsilon",
    Double_epsilon = "Double epsilon"
  ) %>%
  
  tab_options(
    table.font.size = px(13),
    row.striping.include_table_body = TRUE
  )

```
### Visualising the densities of two different predictor variables (elevation and longitude) to see if the epsilon choice has a different effect on the estimates for these two predictors: 

```{r}
plot_df <- bind_rows(
  draws_single %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "Original"),
  
  draws_half %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "Half_epsilon"),
  
  draws_double %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "Double_epsilon")
)

#elevation plot: 
p_elevation <- ggplot(
  plot_df %>% filter(Parameter == "b_Elevation_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Elevation)",
    y = "Density"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
p_elevation

#longitude plot:
p_long <- ggplot(
  plot_df %>% filter(Parameter == "b_Longitude_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Longitude)",
    y = "Density"
  ) +
  theme_minimal()
p_long
```

### Total abundance sensitivity analysis: Epsilon Choice 

```{r}
#C-F contrasts

m_AB_gaussian_FC_single  <- readRDS("m_abundance_CF_gaussian_new.rds")
m_AB_gaussian_FC_dbleps  <- readRDS("model_AB_full_doubleeps.rds")
m_AB_gaussian_FC_hlfeps  <- readRDS("model_AB_full_halfeps.rds")

variables <- names(as_draws_df(m_AB_gaussian_FC_single))
variables[grepl("^b_", variables)]

#look at posterior draws: 
draws_single <- as_draws_df(m_AB_gaussian_FC_single)
draws_half <- as_draws_df(m_AB_gaussian_FC_hlfeps)
draws_double <- as_draws_df(m_AB_gaussian_FC_dbleps)

# parameters to compare
params <- c(
  "b_Intercept",
  "b_Elevation_s",
  "b_Rainfall_mm_s",
  "b_Latitude_s",
  "b_Longitude_s",
  "b_Matrix_meanOpen_land",
  "b_Matrix_meanProduction_forests",
  "b_Matrix_meanSecondaryforest",
  "b_Matrix_meanWater",
  "b_Forest_GroupTropical_Montane"
)


#Summarise all fixed effects across epsilon choices and make a table:
summ_fun <- function(draws, model_name) {
  draws %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    group_by(Parameter) %>%
    summarise(
      Mean  = mean(Value),
      Lower = quantile(Value, 0.025),
      Upper = quantile(Value, 0.975),
      .groups = "drop"
    ) %>%
    mutate(Model = model_name)
}

# create summary tables
tab_single <- summ_fun(draws_single, "Original")
tab_half   <- summ_fun(draws_half, "Half_epsilon")
tab_double <- summ_fun(draws_double, "Double_epsilon")

# combine
summary_table <- bind_rows(
  tab_single,
  tab_half,
  tab_double
) 

# pivot wider for readability
summary_table_wide_ABeps <- summary_table %>%
  
  mutate(
    Estimate = sprintf("%.2f [%.2f, %.2f]",
                       Mean, Lower, Upper)
  ) %>%
  
  select(Parameter, Model, Estimate) %>%
  
  pivot_wider(
    names_from = Model,
    values_from = Estimate
  )

summary_table_wide_ABeps %>%
  mutate(
    Parameter = recode(
      Parameter,
      "b_Intercept" = "Intercept",
      "b_Elevation_s" = "Elevation",
      "b_Rainfall_mm_s" = "Rainfall",
      "b_Latitude_s" = "Latitude",
      "b_Longitude_s" = "Longitude",
      "b_Matrix_meanOpen_land" = "Open land",
      "b_Matrix_meanProduction_forests" = "Production forests",
      "b_Matrix_meanSecondaryforest" = "Secondary forest",
      "b_Matrix_meanWater" = "Water",
      "b_Forest_GroupTropical_Montane" = "Tropical montane"
    )
  ) %>%
  gt(rowname_col = "Parameter") %>%
  tab_header(
    title = "Sensitivity analysis: continuity-correction epsilon"
  ) %>%
  cols_label(
    Original = "Original epsilon",
    Half_epsilon = "Half epsilon",
    Double_epsilon = "Double epsilon"
  ) %>%
  tab_options(
    table.font.size = px(13),
    row.striping.include_table_body = TRUE
  )

```

### In this section, I will be doing a asensitivity analysis on the assigment of 10,000ha to continuous forest, and whether this has a significant effect on the posterior estimates for my model.

```{r}

m_AB_20K <- readRDS("m_abundance_ALL_20K.rds")
m_AB_10K  <- readRDS("m_abundance_interaction_ALL_new.rds")
m_AB_6K  <- readRDS("m_abundance_ALL_6K.rds")

variables <- names(as_draws_df(m_AB_6K))
variables[grepl("^b_", variables)]

#look at posterior draws: 
draws_10K <- as_draws_df(m_AB_10K)
draws_6K <- as_draws_df(m_AB_6K)
draws_20K <- as_draws_df(m_AB_20K)

# parameters to compare
params <- c(
  "b_Intercept",
  "b_ComparisonFrag_vs_Frag",
  "b_log_forest_area_ref_s",
  "b_Elevation_s",
  "b_Rainfall_mm_s",
  "b_Latitude_s",
  "b_Longitude_s",
  "b_Matrix_meanHuman_settlements",
  "b_Matrix_meanOpen_land",
  "b_Matrix_meanProduction_forests",
  "b_Matrix_meanSecondaryforest",
  "b_Matrix_meanWater",
  "b_Forest_GroupTropical_Lowland",
  "b_Forest_GroupTropical_Montane",
  "b_ComparisonFrag_vs_Frag:log_forest_area_ref_s"
)


#Summarise all fixed effects across epsilon choices and make a table:
summ_fun <- function(draws, model_name) {
  draws %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    group_by(Parameter) %>%
    summarise(
      Mean  = mean(Value),
      Lower = quantile(Value, 0.025),
      Upper = quantile(Value, 0.975),
      .groups = "drop"
    ) %>%
    mutate(Model = model_name)
}

# create summary tables
tab_10K <- summ_fun(draws_10K, "10K")
tab_6K   <- summ_fun(draws_6K, "6K")
tab_20K <- summ_fun(draws_20K, "20K")

# combine
summary_table <- bind_rows(
  tab_10K,
  tab_6K,
  tab_20K
) 


summary_table_wide_forestsize <- summary_table %>%
  
  mutate(
    Estimate = sprintf("%.2f [%.2f, %.2f]",
                       Mean, Lower, Upper)
  ) %>%
  
  select(Parameter, Model, Estimate) %>%
  
  pivot_wider(
    names_from = Model,
    values_from = Estimate
  )

summary_table_wide_forestsize %>%
  mutate(
    Parameter = recode(
      Parameter,
      "b_Intercept" = "Intercept",
      "b_ComparisonFrag_vs_Frag" = "F-C comparison",
      "b_ComparisonFrag_vs_Frag:log_forest_area_ref_s" = "F-C x forest area",
      "b_Elevation_s" = "Elevation",
      "b_Rainfall_mm_s" = "Rainfall",
      "b_Latitude_s" = "Latitude",
      "b_Longitude_s" = "Longitude",
      "b_Matrix_meanHuman_settlements" = "Human settlements",
      "b_Matrix_meanOpen_land" = "Open land",
      "b_Matrix_meanProduction_forests" = "Production forests",
      "b_Matrix_meanSecondaryforest" = "Secondary forest",
      "b_Matrix_meanWater" = "Water",
      "b_log_forest_area_ref_s" = "Log forest area",
      "b_Forest_GroupTropical_Lowland" = "Tropical lowland",
      "b_Forest_GroupTropical_Montane" = "Tropical montane"
    )
  ) %>%
  gt(rowname_col = "Parameter") %>%
  tab_header(
    title = "Sensitivity analysis: continuous forest size assumptions"
  ) %>%
  cols_label(
    `10K` = "10,000 ha",
    `6K` = "6,000 ha",
    `20K` = "20,000 ha"
  ) %>%
  tab_options(
    table.font.size = px(13),
    row.striping.include_table_body = TRUE
  )
```
### Plotting the densities of the posterior estimates for these three predictors to visualise how the choice of forest size for continuous forest sites affects the estimates for these predictors:

```{r}
plot_df <- bind_rows(
  draws_10K %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "10K"),
  
  draws_6K %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "6K"),
  
  draws_20K %>%
    select(all_of(params)) %>%
    pivot_longer(
      cols = everything(),
      names_to = "Parameter",
      values_to = "Value"
    ) %>%
    mutate(Model = "20K")
)

#elevation plot: 
p_elevation <- ggplot(
  plot_df %>% filter(Parameter == "b_Elevation_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Elevation)",
    y = "Density"
  ) +
  theme_minimal() 
p_elevation

#longitude plot:
p_long <- ggplot(
  plot_df %>% filter(Parameter == "b_Longitude_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Longitude)",
    y = "Density"
  ) +
  theme_minimal()
p_long

#rainfall plot:
p_rain <- ggplot(
  plot_df %>% filter(Parameter == "b_Rainfall_mm_s"),
  aes(x = Value, fill = Model)
) +
  geom_density(alpha = 0.4) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x = "Posterior estimate (Rainfall)",
    y = "Density"
  ) +
  theme_minimal() 
p_rain
```