### This script analyses species-level bat responses to forest fragmentation using phylogenetically informed Bayesian hierarchical meta-analysis. Responses are partitioned into fragment–continuous forest (F–C) and fragment–fragment (F–F) contrasts, alongside unified models that explicitly test the effects of comparison context. The workflow integrates ecological traits, habitat affinity, diet, conservation status and landscape context while accounting for study-level dependence and phylogenetic non-independence through multilevel random effects and phylogenetic covariance matrices. 

```{r}
# Loading required packages: 
# Load required packages
pkgs <- c(
  "brms", "cmdstanr", "tidybayes", "projpred",
  "dplyr", "tidyr", "forcats", "purrr",
  "ggplot2", "patchwork", "bayesplot", "ggcorrplot",
  "emmeans", "loo", "performance", "corrr",
  "car", "ape", "ggtree", "ggtreeExtra",
  "rstatix", "rotl", "lsr"
)

invisible(lapply(pkgs, library, character.only = TRUE))

#setting working directory: 
setwd("/Users/dillanhoyt/Documents/PhD/Meta Analysis/Statistics/Meta Analysis")

#inserting phylogenetic correlation matrix: 
phylo_cor <- readRDS("/Users/dillanhoyt/Documents/PhD/Meta Analysis/Statistics/Meta Analysis/phylo_cor_matrix.rds")

#Inserting csv: 
meta_species <- read.csv("/Users/dillanhoyt/Documents/PhD/Meta Analysis/Statistics/Meta Analysis/effect_size_species.csv")
```
### Ensuring structure and scaling are in place

```{r}
# Prepare factors
meta_species <- meta_species %>%
  mutate(
    Forest_Group        = as.factor(Forest_Group),
    Matrix_mean         = as.factor(Matrix_mean),
    Conservation_status = as.factor(Conservation_status),
    Diet                = as.factor(Diet),
    Strata              = as.factor(Strata),
    Foraging_strat      = as.factor(Foraging_strat),
    Habitat_affinity_new      = as.factor(Habitat_affinity_new)
    
  )

#scaling all numerical predictors: 

meta_species <- meta_species %>%
  mutate(
    # numeric transforms
    Elevation_s   = as.numeric(scale(Elevation)),
    Rainfall_mm_s = as.numeric(scale(Rainfall_mm)),
    Latitude_s    = as.numeric(scale(Latitude)),
    Longitude_s   = as.numeric(scale(Longitude)),
    FA_length_s   = as.numeric(scale(FA_length)),
    Wing_loading_s  = as.numeric(scale(Wing_loading)),
    Aspect_ratio_s  = as.numeric(scale(Aspect_ratio)),
    Averagemass_g_s  = as.numeric(scale(Averagemass_g)))

#Saving a copy of the meta_species dataframe before I make any changes to it for the models, so I can easily refer back to it if needed.
meta_species1 <- meta_species 

 #Subsetting the species retained and their triat information for large database: 
meta_species1_traits <- meta_species1 %>% dplyr::select(Species,Averagemass_g,FA_length, Diet, Trophic_level,Aspect_ratio, Wing_loading, Habitat_affinity_new, Strata, Foraging_strat)

#One species per row
traits_unique <- meta_species1_traits %>%
  distinct(Species, .keep_all = TRUE)

# Write trait csv:
write.csv(traits_unique, "GlobalBatTraits.csv")

```

# Match species to Open Tree of Life, build a phylogenetic correlation matrix,and prune the dataset to taxa with valid phylogenetic placements.

```{r}
###  choose input source: meta_species1 if present, otherwise fallback to meta_species

meta_species_source <- meta_species
colnames(meta_species_source)
#  helper cleaners
canon <- function(x) {
  x2 <- ifelse(is.na(x), NA_character_, as.character(x))
  x2 <- trimws(x2)
  x2 <- gsub("_", " ", x2)
  x2 <- gsub("[^a-zA-Z ]+", " ", x2)
  x2 <- gsub("\\s+", " ", x2)
  tolower(trimws(x2))
}
clean_species_name <- function(x) {
  y <- as.character(x)
  y <- trimws(y)
  y <- gsub("_ott\\d+$", "", y)            # drop trailing _ott##### if present
  y <- gsub("_", " ", y)                   # underscores -> spaces
  y <- iconv(y, from = "", to = "ASCII//TRANSLIT") # transliterate accents
  y <- gsub("\\b(sp|spp|cf)\\b\\.?","", y, ignore.case = TRUE)
  y <- gsub("\\(.*?\\)", "", y)            # remove parentheses
  y <- gsub("[\"'`]", "", y)
  y <- gsub("[^A-Za-z0-9\\s-]", " ", y)    # strip stray punctuation
  y <- gsub("\\s+", " ", y)
  tolower(trimws(y))
}


# unique cleaned species vector to feed TNRS
species_raw <- meta_species_source %>% 
  mutate(Species_orig = as.character(Species)) %>%
  distinct(Species_orig) %>%
  pull(Species_orig) %>% as.character()

species_clean <- unique(vapply(species_raw, clean_species_name, character(1), USE.NAMES = FALSE))

# run TNRS (single run)
tnrs_matches <- tnrs_match_names(species_clean)
tnrs_df <- as_tibble(tnrs_matches) %>%
  mutate(search_string = as.character(search_string),
         unique_name   = as.character(unique_name),
         ott_id        = suppressWarnings(as.integer(ott_id)))


#  build ott_lookup
ott_lookup <- tnrs_df %>%
  mutate(original_name = trimws(tolower(search_string)),
         Species_phylo  = ifelse(!is.na(unique_name) & !is.na(ott_id),
                                 paste0(gsub(" ", "_", unique_name), "_ott", ott_id),
                                 NA_character_)) %>%
  select(original_name, Species_phylo, ott_id) %>%
  distinct(original_name, .keep_all = TRUE)


# build tree & phylogeny matrix if ott ids exist
ott_ids <- na.omit(unique(ott_lookup$ott_id))

tree <- tol_induced_subtree(ott_ids = ott_ids)
tree_bl <- ape::compute.brlen(tree, method = "Grafen")
phylo_cor_full <- ape::vcv.phylo(tree_bl, corr = TRUE)


# build Species_phylo_use by precedence:
# existing Species_phylo_use -> explicit Species_phylo column -> ott-derived Species_phylo
# join TNRS results back onto the source data
df <- meta_species_source %>%
  mutate(Species_lower = clean_species_name(as.character(Species))) %>%
  left_join(
    ott_lookup %>% rename(Species_lower = original_name),
    by = "Species_lower"
  ) %>%
  mutate(
    Species_phylo_use = as.character(Species_phylo)
  )

# final prune: keep only rows whose Species_phylo_use is present in phylo_cor_full
present <- rownames(phylo_cor_full)
df_final <- df %>%
  filter(!is.na(Species_phylo_use) & Species_phylo_use %in% present) %>%
  mutate(Species = factor(Species_phylo_use, levels = unique(Species_phylo_use))) %>%
  droplevels()

# create phylo_cor pruned to df_final species order
species_order <- levels(df_final$Species)
phylo_cor <- phylo_cor_full[species_order, species_order, drop = FALSE]

meta_species <- df_final


```

# Compute sampling variance for each lnRR using: vi = (1 / nt) + (1 / nc), where nt = focal sample size and nc = comparison sample size. Missing or zero sample sizes are replaced with 0.5,and a minimum variance floor of 0.2 is applied. The square root of vi is used as the observation-level standard error (se_obs).
```{r}
# conservative variance floor
vi_floor <- 0.2

meta_species <- meta_species %>%
  mutate(
    # focal replication
    nt = as.numeric(Sample_size_sum),
    
    # partner (continuous pooled) replication
    nc = as.numeric(Cont_Sample_size_sum),
    
    # safe fallbacks
    nt_fix = ifelse(is.na(nt) | nt <= 0, 0.5, nt),
    nc_fix = ifelse(is.na(nc) | nc <= 0, 0.5, nc),
    
    # sampling variance
    vi_raw = 1/nt_fix + 1/nc_fix,
    vi     = pmax(vi_raw, vi_floor),
    
    # measurement error SD
    se_obs = sqrt(vi)
  )

```
### Separating F-C contrasts from the larger dataframe
```{r}
meta_species_FC <- meta_species %>%
  # keep only fragment–continuous contrasts
  filter(Comparison == "Frag_vs_Cont") %>%
  
  # ensure lnRR is numeric
  mutate(lnRR = as.numeric(lnRR)) %>%
  
  # keep only valid effect sizes and variances
  filter(
    !is.na(lnRR), is.finite(lnRR),
    !is.na(vi),   is.finite(vi), vi > 0
  ) %>%
  
  # ensure modelling IDs are factors
  mutate(
    Paper_ID = factor(Paper_ID),
    Species  = factor(Species)
  ) %>%
  droplevels()
```
### Checking collinearity between numerical predictors in the F-C dataset
```{r}
# numeric columns only
num_vars <- meta_species_FC %>%
  dplyr::select(
    Elevation_s, Longitude_s, Latitude_s, Rainfall_mm_s,
    Wing_loading_s,
    FA_length_s, Aspect_ratio_s, Averagemass_g_s
  )

# correlation matrix (pairwise complete observations)
cor_mat <- cor(num_vars, use = "pairwise.complete.obs")

#plot
ggcorrplot(cor_mat, hc.order = TRUE, lab = TRUE, lab_size = 3)

#running a simple linear model for both predictors of interest to compare their fit to the data and decide which one to include in the models, because FA length and body mass are moderately correlated (r = 0.6) and I want to avoid issues with collinearity in the models.:
lm_FA <- lm(lnRR ~ FA_length_s, data = meta_species_FC)
lm_BM <- lm(lnRR ~ Averagemass_g_s, data = meta_species_FC)

# Compare by AIC value:
AIC(lm_FA, lm_BM) #choosing  bodymass over FA length based on lower AIC and interpretation by wider audience
```

# Fit a phylogenetically informed Bayesian hierarchical meta-analysis using a Gaussian error distribution to test whether body mass, conservation status, habitat affinity, and diet predict species-level responses to forest fragmentation (lnRR) in fragment–continuous forest comparisons. Set reference levels as follows: habitat affinity (clutter), diet (nectarivore), conservation status (least concern).

```{r}
#Reading in the model fitted with a Gaussian error distribution for the fragment–continuous dataset:    
m_batTraits_species.gaussian_CF <- readRDS("m_batTraits_species.gaussian_CF_NEW.rds")

summary(m_batTraits_species.gaussian_CF); pp_check(m_batTraits_species.gaussian_CF)

```
### Fit a phylogenetically informed Bayesian hierarchical meta-analysis using a Student's t error distribution to test whether body mass, conservation status, habitat affinity, and diet predict species-level responses to forest fragmentation (lnRR) in fragment–continuous forest comparisons.

```{r}
#Reading in the model fitted with a Student's t error distribution for the fragment–continuous dataset:
m_batTraits_species.student_CF <- readRDS("m_batTraits_species.student_CF_NEW.rds")

summary(m_batTraits_species.student_CF); pp_check(m_batTraits_species.student_CF)

```
### Conducting a LOO comparison of the Gaussian and Student's t models for the fragment–continuous dataset to assess which error distribution provides a better fit to the data.
```{r}
loo_gaussian_CF <- loo(m_batTraits_species.gaussian_CF)
loo_student_CF  <- loo(m_batTraits_species.student_CF)
model_comparison <- loo_compare(loo_gaussian_CF, loo_student_CF)
print(model_comparison) 

# Assess estimated degrees of freedom (nu)
VarCorr(m_batTraits_species.gaussian_CF)
posterior_summary(m_batTraits_species.student_CF, variable = "nu") #extremely heavy-tailed distribution, siggesting that Gaussian is the better fit for the data.

```
###Assessing how much variance is explained by the phylogenetic random effect in the best-fitting model for the fragment–continuous dataset by calculating the phylogenetic heritability (H²) using the variance components from the model summary.
```{r}
# Extract variance components from the best-fitting model (assuming it's the Student's t model)
vc <- VarCorr(m_batTraits_species.student_CF)

# Extract variances
var_phylo  <- vc$Species$sd[1]^2
var_paper  <- vc$Paper_ID$sd[1]^2
var_resid  <- vc$residual__$sd^2

# Phylogenetic heritability
H2_phylo <- var_phylo / (var_phylo + var_paper + var_resid)
H2_phylo

```
### Looking at additional traits and predictors in the dataset, and how they relate to species responses to fragmentation in the fragment–continuous dataset.

```{r}
#Strata and aspect ratio models for fragment–continuous dataset:
m_batTraits_species.AR_Strata_CF <- readRDS("m_batTraits_species.student_CF_AspectRatio_Strata.rds")
summary(m_batTraits_species.AR_Strata_CF) #No real patterns emerging here

#Strata and wing loading models for fragment–continuous dataset:
m_batTraits_species.strata_wl_CF <- readRDS("m_batTraits_species.strata_wl_CF.rds")
summary(m_batTraits_species.strata_wl_CF) #No real patterns emerging here

#Foragin strategy and habitat affinity models for fragment–continuous dataset:
m_batTraits_species.forag_habitat_CF <- readRDS("m_batTraits_species.forag_habitat_CF.rds")
summary(m_batTraits_species.forag_habitat_CF) #No real patterns emerging here

```

### Looking at interactions terms between different traits (i.e., body mass and diet) with C-F contrasts

```{r}
# Body mass and diet interaction model for fragment–continuous dataset:
m_batTraits_species.student_CF_bodymassDiet <- readRDS("m_batTraits_species.student_CF_bodymassDiet_new")
summary(m_batTraits_species.student_CF_bodymassDiet) #No real patterns emerging here

# Body mass and habitat affinity interaction model for fragment–continuous dataset:
m_batTraits_species.bodymass_HabitatInt_CF <- readRDS("m_batTraits_species.student_CF_StrataDiet")
summary(m_batTraits_species.bodymass_HabitatInt_CF) #No real patterns emerging here

```
### Subsetting original species dataframe to fragment - fragment (F-F) contrasts to run similar models to the above for this subset of the data.

```{r}
meta_species_FF <- meta_species %>%
  filter(Comparison == "Frag_vs_Frag") %>%
  mutate(
    size_small = Fragment_size_ha,   # the smaller of the pair 
    log_size_small = log(size_small)
  ) %>%
  filter(is.finite(log_size_small)) #creating a logged number that is the smallest fragment of the two being compared

#scaling log_size_small: 
meta_species_FF <- meta_species_FF %>%
  mutate(log_size_small_s = as.numeric(scale(log_size_small)))

```
# Fit a phylogenetically informed Bayesian hierarchical meta-analysis using a Gaussian error distribution to test whether body mass, conservation status, habitat affinity, and diet predict species-level responses to forest fragmentation (lnRR) in fragment–fragment comparisons. Set reference levels as follows: habitat affinity (clutter), diet (nectarivore), conservation status (least concern)

```{r}
#Loading in the model fitted with a Gaussian error distribution for the fragment–fragment dataset:
m_batTraits_species.gaussian_FF_new <- readRDS("m_batTraits_species.gaussian_FF_new.rds")
summary(m_batTraits_species.gaussian_FF_new); pp_check(m_batTraits_species.gaussian_FF_new) 


```
### Combining all contrasts (F-C and F-F) into a single dataset to run unified models that explicitly test the effects of comparison context (i.e., whether the contrast is F-C or F-F) on species responses to fragmentation, and how this interacts with traits such as body mass, diet and habitat affinity.

```{r}

cf_area_ha <- 10000  # continuous forest reference

meta_species_All <- meta_species %>%
  
  # Keep only relevant comparisons 
  filter(Comparison %in% c("Frag_vs_Cont", "Frag_vs_Frag", "Cont_vs_Frag")) %>%
  
  mutate(
    .orig_comp = Comparison,
    
    # Standardise labels 
    Comparison = case_when(
      .orig_comp %in% c("Frag_vs_Frag", "Frag_vs_Cont") ~ .orig_comp,
      .orig_comp == "Cont_vs_Frag" ~ "Frag_vs_Cont",
      TRUE ~ NA_character_
    ),
    
    # Ensure numeric
    Fragment_size_ha         = as.numeric(Fragment_size_ha),
    Partner_Fragment_size_ha = as.numeric(Partner_Fragment_size_ha)
  ) %>%
  
  # keep valid comparisons
  filter(!is.na(Comparison)) %>%
  
  # Define reference (less fragmented = larger area)
  mutate(
    forest_area_ref_ha = case_when(
      Comparison == "Frag_vs_Frag" ~ 
        pmax(Fragment_size_ha, Partner_Fragment_size_ha, na.rm = TRUE),
      
      Comparison == "Frag_vs_Cont" ~ 
        cf_area_ha,
      
      TRUE ~ NA_real_
    ),
    
    forest_area_ref_ha = if_else(is.finite(forest_area_ref_ha),
                                 forest_area_ref_ha, NA_real_)
  ) %>%
  
  # Log + scale
  mutate(
    log_forest_area_ref   = log(forest_area_ref_ha),
    log_forest_area_ref_s = as.numeric(scale(log_forest_area_ref))
  ) %>%
  
  # 5. Drop missing
  filter(is.finite(log_forest_area_ref)) %>%
  
  # 6. create smaller fragment covariate
  mutate(
    log_size_small = case_when(
      Comparison == "Frag_vs_Frag" ~ 
        log(pmin(Fragment_size_ha, Partner_Fragment_size_ha)),
      
      Comparison == "Frag_vs_Cont" ~ 
        log(Fragment_size_ha),
      
      TRUE ~ NA_real_
    ),
    
    log_size_small_s = as.numeric(scale(log_size_small))
  ) %>%
  
  select(-.orig_comp)

```
# Fit a phylogenetically informed Bayesian hierarchical model with Contrast type as fixed effect: 

```{r}
m_batTraits_species.gaussian_ALL_NEW <- readRDS("m_batTraits_species.gaussian_ALL_NEW.rds")
summary(m_batTraits_species.gaussian_ALL_NEW); pp_check(m_batTraits_species.gaussian_ALL_NEW)
```
### Fit model to allow species to differ in their responses between fragment–continuous and fragment–fragment contrasts by including a species-level random slope for Comparison nested within the phylogenetic covariance structure.

```{r}
m_stable_uncor1 <- readRDS("m_batTraits_species.uncorr1_gaussian.rds")
summary(m_stable_uncor1); pp_check(m_stable_uncor1)

```
### Funnel plot to check for publication bias in the fragment–continuous dataset by plotting the effect sizes (lnRR) against their standard errors (se_obs) and adding a reference line at lnRR = 0.

```{r}
ggplot(meta_species_FC, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()

ggplot(meta_species_All, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()

ggplot(meta_species_FF, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()
```