---
title: "Acoustic-Based Species Level Meta Analysis "
subtitle: "Analyzing the effect of fragmentation on species-level activity from bioacoustic data by investigating how traits such as body mass, diet, habitat affinity and conservation status influence species responses to fragmentation"
author: "Dillan Hoyt"
execute: 
  eval: true
  echo: false
  message: false
editor: visual
toc: true
toc-depth: 3
toc-title: Contents
---
### This script analyses acoustic-based species-level bat responses to forest fragmentation using phylogenetically informed Bayesian hierarchical meta-analysis. Responses are partitioned into fragment–continuous forest (F–C) and fragment–fragment (F–F) contrasts, alongside unified models that explicitly test the effects of comparison context. The workflow integrates ecological traits, habitat affinity, diet, conservation status and landscape context while accounting for study-level dependence and phylogenetic non-independence through multilevel random effects and phylogenetic covariance matrices. 
### Load necessary libraries and data:

```{r}
#Load libraries
libs <- c(
  "projpred", "lsr", "tidybayes", "brms",
  "dplyr", "tidyr", "ggplot2",
  "bayesplot", "patchwork", "emmeans",
  "forcats", "loo", "performance",
  "corrr", "ggcorrplot", "car",
  "ape", "ggtree", "ggtreeExtra", "rotl",
  "rstatix", "stringr", "readr", "tibble"
)

invisible(lapply(libs, library, character.only = TRUE))

#insert csv:
meta_ACT <- read.csv("effect_size_totalactivity.csv")

#Prepare factors and scale variables:
meta_ACT <- meta_ACT %>%
  mutate(
    Forest_Group        = as.factor(Forest_Group),
    Matrix_mean         = as.factor(Matrix_mean),
    Conservation_status = as.factor(Conservation_status),
    Diet                = as.factor(Diet),
    Strata              = as.factor(Strata),
    Foraging_strat      = as.factor(Foraging_strat),
    Habitat_affinity_new= as.factor(Habitat_affinity_new)
  )

meta_ACT <- meta_ACT %>%
  mutate(
    # numeric transforms
    Elevation_s   = as.numeric(scale(Elevation)),
    Rainfall_mm_s = as.numeric(scale(Rainfall_mm)),
    Latitude_s    = as.numeric(scale(Latitude)),
    Longitude_s   = as.numeric(scale(Longitude)),
    FA_length_s   = as.numeric(scale(FA_length)),
    Wing_loading_s  = as.numeric(scale(Wing_loading)),
    Aspect_ratio_s  = as.numeric(scale(Aspect_ratio)),
    Averagemass_g_s  = as.numeric(scale(Averagemass_g)))

```
#Removing species groups and sonotypes and then matching the remaining species to a phylogeny:

```{r}
# strict but reasonably permissive binomial pattern:
binomial_rgx <- "^[A-Z][A-Za-z'`-]* [a-z][A-Za-z'`-]*$"

# prepare species column (trim whitespace)
meta_ACT <- meta_ACT %>%
  mutate(Species_clean = str_trim(Species))

# logical mask: TRUE when looks like a single binomial
is_binomial <- str_detect(meta_ACT$Species_clean, binomial_rgx)

# subset to only binomials
meta_ACT_binomial <- meta_ACT %>% filter(is_binomial)


meta_ACT<- meta_ACT_binomial

canon <- function(x) {
  x2 <- ifelse(is.na(x), NA_character_, as.character(x))
  x2 <- trimws(x2)
  x2 <- gsub("_", " ", x2)
  x2 <- gsub("[^a-zA-Z ]+", " ", x2)
  x2 <- gsub("\\s+", " ", x2)
  tolower(trimws(x2))
}
clean_species_name <- function(x) {
  y <- as.character(x)
  y <- trimws(y)
  y <- gsub("_ott\\d+$", "", y)
  y <- gsub("_", " ", y)
  y <- iconv(y, from = "", to = "ASCII//TRANSLIT")
  y <- gsub("\\b(sp|spp|cf)\\b\\.?","", y, ignore.case = TRUE)
  y <- gsub("\\(.*?\\)", "", y)
  y <- gsub("[\"'`]", "", y)
  y <- gsub("[^A-Za-z0-9\\s-]", " ", y)
  y <- gsub("\\s+", " ", y)
  tolower(trimws(y))
}

#Prepare species list for matching to phylogeny:
species_raw <- meta_ACT %>%
  mutate(Species_orig = as.character(Species)) %>%
  distinct(Species_orig) %>%
  pull(Species_orig) %>%
  as.character()

species_clean <- unique(vapply(species_raw, clean_species_name, character(1), USE.NAMES = FALSE))

tnrs_matches <- tnrs_match_names(species_clean)
tnrs_df <- as_tibble(tnrs_matches) %>%
  mutate(search_string = as.character(search_string),
         unique_name   = as.character(unique_name),
         ott_id        = suppressWarnings(as.integer(ott_id)))

#Build ott lookup
ott_lookup <- tnrs_df %>%
  mutate(original_name = trimws(tolower(search_string)),
         Species_phylo  = ifelse(!is.na(unique_name) & !is.na(ott_id),
                                 paste0(gsub(" ", "_", unique_name), "_ott", ott_id),
                                 NA_character_)) %>%
  select(original_name, Species_phylo, ott_id) %>%
  distinct(original_name, .keep_all = TRUE)

#build tree & phylogeny matrix
ott_ids <- na.omit(unique(ott_lookup$ott_id))
tree <- tol_induced_subtree(ott_ids = ott_ids)
tree_bl <- ape::compute.brlen(tree, method = "Grafen")
phylo_cor_full <- ape::vcv.phylo(tree_bl, corr = TRUE)

#join ott_lookup back into meta_ACT
df <- meta_ACT %>%
  mutate(Species_lower = clean_species_name(as.character(Species))) %>%
  left_join(ott_lookup %>% rename(Species_lower = original_name),
            by = "Species_lower") %>%
  mutate(Species_phylo = ifelse(is.na(Species_phylo), NA_character_, as.character(Species_phylo)),
         ott_id = as.integer(ott_id))

present <- rownames(phylo_cor_full)

df_final <- df %>%
  filter(!is.na(Species_phylo) & Species_phylo %in% present) %>%
  mutate(Species = factor(Species_phylo, levels = unique(Species_phylo))) %>%
  droplevels()

# pruned phylogeny correlation matrix
species_order <- levels(df_final$Species)
phylo_cor <- phylo_cor_full[species_order, species_order, drop = FALSE]

meta_ACT <- df_final

```

### Adding a weight and missing habitat affinity category for species with missing habitat affinity data:
```{r}
vi_floor <- 0.2

meta_ACT <- meta_ACT %>%
  mutate(
    # focal replication
    nt = as.numeric(Sample_size_sum),
    
    # partner (continuous pooled) replication
    nc = as.numeric(Cont_Sample_size_sum),
    
    # safe fallbacks
    nt_fix = ifelse(is.na(nt) | nt <= 0, 0.5, nt),
    nc_fix = ifelse(is.na(nc) | nc <= 0, 0.5, nc),
    
    # sampling variance
    vi_raw = 1/nt_fix + 1/nc_fix,
    vi     = pmax(vi_raw, vi_floor),
    
    # measurement error SD
    se_obs = sqrt(vi)
  )

  meta_ACT <- meta_ACT %>%
  mutate(
    Habitat_affinity_new = case_when(
      
      Species %in% c("Noctilio_leporinus_ott604404",
                     "Noctilio_albiventris_ott604407") ~ "Water-adapted",
      TRUE ~ Habitat_affinity_new
    )
  )

meta_ACT <- meta_ACT %>%
  mutate(
    Habitat_affinity_new = case_when(
      
      Species %in% c("Rhinolophus_beddomei_ott994640",
                     "Rhinolophus_indorouxii_ott7067837",
                     "Rhinolophus_lepidus_ott217411",
                     "Rhinolophus_rouxii_ott1047994",
                     "Hipposideros_pomona_ott905428",
                     "Hipposideros_commersoni_ott156174",
                     "Scotomanes_ornatus_ott759645") ~ "Clutter-adapted",
      TRUE ~ Habitat_affinity_new
    )
  )
      
meta_ACT <- meta_ACT %>%
  mutate(
    Habitat_affinity_new = case_when(
      
      Species %in% c("Miniopterus_fuliginosus_ott1011476",
                     "Miniopterus_pusillus_ott61867") ~ "Open-adapted",
      TRUE ~ Habitat_affinity_new
    )
  )

meta_ACT <- meta_ACT %>%
  mutate(
    Habitat_affinity_new = case_when(
      
      Species %in% c("Pipistrellus_ceylonicus_ott743848",
                     "Hesperoptenus_tickelli_ott445742",
                     "Peropteryx_trinitatis_ott432533",
                     "Saccopteryx_canescens_ott232381",
                     "Perimyotis_subflavus_ott977400",
                     "Emballonura_atrata_ott156176") ~ "Edge-adapted",
      TRUE ~ Habitat_affinity_new
    )
  )
```
### Adding call type categories for species with missing call type data:

```{r}
meta_ACT <- meta_ACT %>%
  mutate(
    Call_type_clean = case_when(
      is.na(Call_type) ~ NA_character_,
      
      # CF-dominated (any string starting with CF)
      str_detect(Call_type, "^CF") ~ "CF",
      
      # FM.QCF, FM_QCF, FM-QCF -> treat as FM-dominated
      str_detect(Call_type, "^FM") ~ "FM",
      
      # Pure QCF
      Call_type == "QCF" ~ "QCF",
      
      TRUE ~ NA_character_
    )
  )

  meta_ACT <- meta_ACT %>%
  mutate(
    Call_type_clean = case_when(
      
      # CF specialists 
      Species %in% c(
        "Rhinolophus_beddomei_ott994640",
        "Rhinolophus_indorouxii_ott7067837",
        "Rhinolophus_lepidus_ott217411",
        "Rhinolophus_rouxii_ott1047994",
        "Hipposideros_commersoni_ott156174",
        "Hipposideros_pomona_ott905428"
      ) ~ "CF",
      
      # QCF (narrowband open-space molossid style) 
      Species %in% c(
        "Molossus_currentium_ott3614007",
        "Molossus_rufus_ott267993",
        "Promops_centralis_ott301613",
        "Nyctinomops_macrotis_ott14996",      
        "Nyctinomops_laticaudatus_ott14995"   
      ) ~ "QCF",
      
      # FM-dominated 
      Species %in% c(
        "Hesperoptenus_tickelli_ott445742",
        "Miniopterus_fuliginosus_ott1011476",
        "Miniopterus_pusillus_ott61867",
        "Miniopterus_manavi_ott18883",
        "Pipistrellus_ceylonicus_ott743848",
        "Peropteryx_trinitatis_ott432533",
        "Saccopteryx_canescens_ott232381",
        "Emballonura_atrata_ott156176",
        "Perimyotis_subflavus_ott977400",
        "Myotis_albescens_ott353889",
        "Myotis_nigricans_ott307133",
        "Myotis_goudoti_ott31985",
        "Eptesicus_fuscus_ott10737",
        "Lasionycteris_noctivagans_ott401283",
        "Lasiurus_borealis_ott61860",
        "Aeorestes_cinereus_ott369537",
        "Cyttarops_alecto_ott599540",
        "Diclidurus_albus_ott1060477",
        "Dasypterus_ega_ott635119"
      ) ~ "FM",
      
      TRUE ~ Call_type_clean
    )
  )


```

### Looking at F-C contrasts only:
```{r}
meta_ACT_FC <- meta_ACT %>%
  # keep only fragment–continuous contrasts
  filter(Comparison == "Frag_vs_Cont") %>%
  
  # ensure lnRR is numeric
  mutate(lnRR = as.numeric(lnRR)) %>%
  
  # keep only valid effect sizes and variances
  filter(
    !is.na(lnRR), is.finite(lnRR),
    !is.na(vi),   is.finite(vi), vi > 0
  ) %>%
  
  # ensure modelling IDs are factors
  mutate(
    Paper_ID = factor(Paper_ID),
    Species  = factor(Species)
  ) %>%
  droplevels()
```

### Models for F-C contrasts only:

```{r}
#Gaussian model
m_batTraits_species.gaussian_CF_ACT <- readRDS("m_batTraits_species.gaussian_CF_ACT.rds")
summary(m_batTraits_species.gaussian_CF_ACT)

#Student t model
m_batTraits_species.student_CF_ACT <- readRDS("m_batTraits_species.student_CF_ACT.rds")
summary(m_batTraits_species.student_CF_ACT)

mcmc_intervals(as_draws_df(m_batTraits_species.student_CF_ACT), pars = vars(starts_with("b_")),
               prob = 0.95) +
  vline_at(0, linetype = "dashed")

#loo to compare gaussian vs student: 
loo_trait_gaussian <- loo(m_batTraits_species.gaussian_CF_ACT)
loo_trait_student <- loo(m_batTraits_species.student_CF_ACT)
loo_compare(loo_trait_gaussian, loo_trait_student)

```

### Subsetting for fragment fragment contrasts only:

```{r}
meta_ACT_FF <- meta_ACT %>%
  filter(Comparison == "Frag_vs_Frag") %>%
  mutate(
    size_small = Fragment_size_ha,   # the smaller of the pair (from your construction)
    log_size_small = log(size_small)
  ) %>%
  filter(is.finite(log_size_small)) #creating a logged number that is the smallest fragment of the two being compared, because the size of the smaller fragment is the variable that changes between comparisons — the large fragment just acts as a reference.

#scaling log_size_small: 
meta_ACT_FF <- meta_ACT_FF %>%
  mutate(log_size_small_s = as.numeric(scale(log_size_small)))
```
### F-F models:

```{r}
m_batTraits_species.student_FF_ACT <- readRDS("m_batTraits_species.student_FF_ACT.rds")
summary(m_batTraits_species.student_FF_ACT)

mcmc_intervals(as_draws_df(m_batTraits_species.student_FF_ACT), pars = vars(starts_with("b_")),
               prob = 0.95) +
  vline_at(0, linetype = "dashed")

```

### Combining F-C and F-F contrasts into a single model with an explicit comparison context variable:

```{r}
force_cf <- FALSE  

meta_ACT_All <- meta_ACT %>%
  # 1. canonicalise contrasts & flip lnRR for Cont_vs_Frag
  mutate(
    .orig_comp = Comparison,
    lnRR = if_else(.orig_comp == "Cont_vs_Frag", -lnRR, lnRR),
    Comparison = case_when(
      .orig_comp %in% c("Frag_vs_Frag", "Frag_vs_Cont") ~ .orig_comp,
      .orig_comp == "Cont_vs_Frag"                       ~ "Frag_vs_Cont",
      TRUE                                               ~ NA_character_
    ),
    Fragment_size_ha         = as.numeric(Fragment_size_ha),
    Partner_Fragment_size_ha = as.numeric(Partner_Fragment_size_ha)
  ) %>%
  
  # keep only recognised contrasts
  filter(!is.na(Comparison)) %>%
  
  # 2. For Frag_vs_Cont rows where the recorded fragment (numerator) has NO size,
  #    treat the continuous forest as numerator: flip lnRR and assign CF = 10000.
  mutate(
    frag_missing = (Comparison == "Frag_vs_Cont" & Treatment == "Frag" & is.na(Fragment_size_ha)),
    
    # flip lnRR and Treatment where fragment side is missing (so CF becomes numerator)
    lnRR = if_else(frag_missing, -lnRR, lnRR),
    Treatment = if_else(frag_missing, "Cont", Treatment),
    
    # set Partner_Fragment_size_ha to 10000 for the continuous partner that will be used as numerator
    Partner_Fragment_size_ha = case_when(
      frag_missing ~ 10000,
      force_cf == TRUE & Comparison == "Frag_vs_Cont" & Treatment == "Cont" ~ 10000,
      identical(force_cf, "missing_or_zero") & Comparison == "Frag_vs_Cont" & Treatment == "Cont" & (is.na(Partner_Fragment_size_ha) | Partner_Fragment_size_ha == 0) ~ 10000,
      TRUE ~ Partner_Fragment_size_ha
    )
  ) %>%
  
  # 3. compute numerator area depending on canonicalised Comparison/Treatment
  mutate(
    area_num_ha = case_when(
      Comparison == "Frag_vs_Frag"                            ~ Fragment_size_ha,                      # smaller fragment stored in Fragment_size_ha
      Comparison == "Frag_vs_Cont" & Treatment == "Frag"      ~ Fragment_size_ha,                      # fragment is numerator
      Comparison == "Frag_vs_Cont" & Treatment == "Cont"      ~ Partner_Fragment_size_ha,              # partner (CF) is numerator (possibly set to 10000 above)
      TRUE                                                   ~ NA_real_
    )
  ) %>%
  
  # 4. log-transform and scale
  mutate(
    log_area_num   = as.numeric(log(area_num_ha)),
    log_area_num_s = as.numeric(scale(log_area_num))
  ) %>%
  
  # 5. drop rows w/ missing area and remove helpers
  filter(is.finite(log_area_num)) %>%
  select(-.orig_comp, -frag_missing)

  #making sure that the smaller frag is the numerator
meta_ACT_All <- meta_ACT_All %>%
  mutate(
    # ensure for Frag_vs_Frag the numerator area is the smaller of the two recorded fragment sizes
    area_num_ha = case_when(
      Comparison == "Frag_vs_Frag" ~ pmin(Fragment_size_ha, Partner_Fragment_size_ha, na.rm = TRUE),
      TRUE                          ~ area_num_ha
    ),
    log_area_num   = as.numeric(log(area_num_ha)),
    log_area_num_s = as.numeric(scale(log_area_num))
  ) %>%
  filter(is.finite(log_area_num))

  # 1) Create s_obs (prefer existing se_obs, otherwise compute from vi)
meta_ACT_All <- meta_ACT_All %>%
  mutate(
    se_obs = if ("se_obs" %in% names(.)) se_obs else NA_real_,   # keep if exists
    vi     = if ("vi"     %in% names(.)) vi     else NA_real_,   # keep if exists
    # create s_obs: prefer se_obs, otherwise sqrt(vi), otherwise NA
    s_obs = case_when(
      !is.na(se_obs) ~ as.numeric(se_obs),
      !is.na(vi)     ~ as.numeric(sqrt(vi)),
      TRUE           ~ NA_real_
    )
  )

m_batTraits_species.student_ACT_ALL <- readRDS("m_batTraits_species.student_ACT_ALL.rds")
summary(m_batTraits_species.student_ACT_ALL)

mcmc_intervals(as_draws_df(m_batTraits_species.student_ACT_ALL), pars = vars(starts_with("b_")),
               prob = 0.95) +
  vline_at(0, linetype = "dashed")

```

```

```