---
title: "Species Richness - Meta Analysis "
subtitle: "Analyzing the effect of fragmentation on species richness from capture data and how this is influenced by landscape variables"
author: "Dillan Hoyt"
execute: 
  eval: true
  echo: false
  message: false
editor: visual
toc: true
toc-depth: 3
toc-title: Contents
---
```{r}
libs <- c(
  "brms", "projpred", "tidybayes", "bayesplot", "emmeans",
  "posterior", "performance", "loo", "car", "dplyr",
  "tidyr", "tibble", "forcats", "ggplot2", "ggdist",
  "ggcorrplot", "scales", "patchwork", "corrr"
)

invisible(lapply(libs, library, character.only = TRUE))

#insert csv: 
meta_SR <- read.csv("effect_size_totalSR.csv")

#removing any edge sites 
meta_SR <- meta_SR %>%
  dplyr::filter(!grepl("Edge", UniqueFragID))

#removing fragments greater than 10,000ha 
meta_species <-subset(meta_SR,Paper_ID!= 34)
#removing repetitions (the same data was included twice by accident)
meta_species <-subset(meta_SR,Paper_ID!= 52)
#removing fragment over 10,000
meta_SR %>%
  filter(Fragment_size_ha > 11000) %>%
  select(Paper_ID, UniqueFragID, Fragment_size_ha) %>%
  distinct() 
```
### Ensuring structure is correct: 

```{r}
# Conservative floor for tiny variances
vi_floor <- 0.2

# Build analysis data with correct sample-size columns
meta_SR2 <- meta_SR %>%
  mutate(
    # make sure these are numeric
    nt = as.numeric(Sample_size),
    
    # use the correct partner sample size depending on comparison type
    nc = case_when(
      Comparison == "Frag_vs_Cont" ~ as.numeric(Cont_Sample_size_sum),
      Comparison == "Frag_vs_Frag" ~ as.numeric(Partner_Sample_size_sum),
      TRUE ~ NA_real_
    ),
    
    # replace missing / zero values with a small placeholder
    nt_fix = if_else(is.na(nt) | nt <= 0, 0.5, nt),
    nc_fix = if_else(is.na(nc) | nc <= 0, 0.5, nc),
    
    # sampling variance
    vi_raw = 1 / nt_fix + 1 / nc_fix,
    vi = pmax(vi_raw, vi_floor),
    se_obs = sqrt(vi)
  ) %>%
  filter(is.finite(vi) & vi > 0)

# Create a consistently oriented effect size for plotting
meta_SR2 <- meta_SR2 %>%
  mutate(
    Comparison = case_when(
      is.na(Comparison) & !is.na(Treatment) & Treatment == "Frag" ~ "Frag_vs_Cont",
      TRUE ~ Comparison
    ),
    lnRR_for_plot = case_when(
      Comparison == "Frag_vs_Cont" ~ lnRR,
      Comparison == "Cont_vs_Frag" ~ -lnRR,
      Comparison == "Frag_vs_Frag" ~ -lnRR,
      TRUE ~ lnRR
    )
  )

# Final analysis dataset
meta_SR <- meta_SR2

#scaling everything: 
meta_SR <- meta_SR %>%
  mutate(
    Elevation_s      = as.numeric(scale(Elevation)),
    Rainfall_mm_s    = as.numeric(scale(Rainfall_mm)),
    Latitude_s       = as.numeric(scale(Latitude)),
    Longitude_s      = as.numeric(scale(Longitude)),
    Fragment_size_ha_s = as.numeric(scale(Fragment_size_ha))
  )

  #before running models
options(mc.cores = parallel::detectCores())
```
### Subsetting data to only look at continuous forest - fragment comparisons and then running a Bayesian hierarchical model to look at the effect of landscape variables on species richness response to fragmentation:

```{r}
meta_SR_FC <- meta_SR %>%
  # keep only Frag vs Cont contrasts
  filter(Comparison %in% c("Frag_vs_Cont", "Cont_vs_Frag")) %>%
  # orient so all effects mean "fragment relative to continuous"
  mutate(
    lnRR_dir = ifelse(Treatment == "Cont", -lnRR, lnRR)
  )

m_SR_gaussian_FC  <- readRDS("model_sr_full_gaussian.rds")
summary(m_SR_gaussian_FC); pp_check(m_SR_gaussian_FC) 

#student t
m_SR_studentt_FC  <- readRDS("model_sr_full_student.rds")
summary(m_SR_studentt_FC); pp_check(m_SR_studentt_FC) 

```
### Subsetting data to only look at fragment - fragment comparisons and then running a Bayesian hierarchical model to look at the effect of landscape variables on species richness response to fragmentation:

```{r}
meta_SR_FF <- meta_SR %>%
filter(Comparison == "Frag_vs_Frag") %>%
  mutate(
    size_small = Fragment_size_ha,   # the smaller of the pair (from your construction)
    log_size_small = log(size_small)
  ) %>%
  filter(is.finite(log_size_small)) #creating a logged number that is the smallest fragment of the two being compared, because the size of the smaller fragment is the variable that changes between comparisons — the large fragment just acts as a reference.

#scaling log_size_small: 
 meta_SR_FF <- meta_SR_FF %>%
  mutate(log_size_small_s = as.numeric(scale(log_size_small)))

#gaussian
m_SR_gaussian_FF  <- readRDS("m_SR_FF_new.rds")
summary(m_SR_gaussian_FF); pp_check(m_SR_gaussian_FF) 

```
### Assigning continuous forest the value of 10,000ha and then running a model with all data to look at the effect of fragment size on species richness response to fragmentation:

```{r}
#applying a 10,000 ha cap on continuous forests
 cf_area_ha <- 10000  # conventional area assigned to continuous forest
 
 meta_SR_All <- meta_SR %>%
   mutate(
     .orig_comp = Comparison,
     
     # Standardise comparison labels only.
     # No lnRR sign changes are made here.
     Comparison = case_when(
       .orig_comp %in% c("Frag_vs_Frag", "Frag_vs_Cont") ~ .orig_comp,
       .orig_comp == "Cont_vs_Frag" ~ "Frag_vs_Cont",
       TRUE ~ NA_character_
     ),
     
     # Make area columns numeric
     Fragment_size_ha = as.numeric(Fragment_size_ha),
     Partner_Fragment_size_ha = as.numeric(Partner_Fragment_size_ha),
     Cont_Fragment_size_ha = as.numeric(Cont_Fragment_size_ha)
   ) %>%
   
   # Keep only recognised contrast types
   filter(!is.na(Comparison)) %>%
   
   mutate(
     # Define the moderator as the area of the less fragmented habitat
     # F-F: larger of the two fragment areas
     # C-F: continuous forest = 10,000 ha
     forest_area_ref_ha = case_when(
       Comparison == "Frag_vs_Frag" ~ pmax(Fragment_size_ha, Partner_Fragment_size_ha, na.rm = TRUE),
       Comparison == "Frag_vs_Cont" ~ cf_area_ha,
       TRUE ~ NA_real_
     ),
     
     # Replace non-finite values
     forest_area_ref_ha = if_else(is.finite(forest_area_ref_ha), forest_area_ref_ha, NA_real_),
     
     # Log-transform and scale
     log_forest_area_ref   = log(forest_area_ref_ha),
     log_forest_area_ref_s = as.numeric(scale(log_forest_area_ref))
   ) %>%
   
   # Drop rows that still lack a usable moderator
   filter(is.finite(log_forest_area_ref)) %>%
   
   # Remove helper column
   select(-.orig_comp)
 
 
 meta_SR_All <- meta_SR_All %>%
   mutate(
     log_size_small = case_when(
       Comparison == "Frag_vs_Frag" ~ log(pmin(Fragment_size_ha, Partner_Fragment_size_ha)),
       Comparison == "Frag_vs_Cont" ~ log(Fragment_size_ha),
       TRUE ~ NA_real_
     ),
     log_size_small_s = as.numeric(scale(log_size_small))
   )
```
### Runnign a model with all data to look at the effect of fragment size on species richness response to fragmentation:

```{r}
# Gaussian model using observation SE
m_SR_gaussian_ALL  <- readRDS("m_sr_all_new.rds")
summary(m_SR_gaussian_ALL); pp_check(m_SR_gaussian_ALL) 

#adding an interaction term with area and comparison type 
m_SR_gaussian_ALL_interaction  <- readRDS("m_sr_all_randomeffect_andinteraction_new.rds")
summary(m_SR_gaussian_ALL_interaction); pp_check(m_SR_gaussian_ALL_interaction) 

#Funnel plots for each of my datasets (F-F, F-C and all data) to check for publication bias:
ggplot(meta_SR_FC, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()

ggplot(meta_SR_FF, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()

ggplot(meta_SR_All, aes(x = se_obs, y = lnRR)) +
  geom_point() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(x = "Standard Error (se_obs)", y = "Effect Size (lnRR)") +
  theme_minimal()
```